---
title: "tAIpipe Comprehensive Analytical Report"
subtitle: "Global and Local Translation Optimization Landscape"
author: "tAIpipe Automated Framework"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: flatly
    highlight: tango
  md_document:
    variant: markdown_github
params:
  samples: ""
  dataset: ""
  per_genome_dir: ""
  output_dir: ""
always_allow_html: true
---

#TODO: Brakuje ogolnei wykresow najprostszych, mean_tai vs phylum/lifestyle, delta_ENC vs phylum/lifestyle, mean_GC3s vs phylum/lifestyle, tAI vs obecnosc/brak {domen PFAM, peptydow sygnalowych, LCR, liczebnosci LCR, dlugosci LCR} itd. Najlepiej to dla present/ansent cech zrobic jeden typ wykresu ktory jako argument wybiera kolumne z tej ogromnej tabeli i plotuje ją vs tAI
#TODO: Brakuje testów statystycznych. Warto dodać gdzies na gorze wykresu oznaczenie '>=0.05' lub przeciwne jesli nie wykryto zaleznosci statystycznie isttonyhc
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 7)

library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
library(cowplot)
library(pheatmap)
library(plotly)
library(DT)

# STANDARDIZED SECTIONS PATH DEFINITIONS
sec0_dir <- file.path(params$output_dir, "summary_report_files", "sec0_quality_control")
sec1_dir <- file.path(params$output_dir, "summary_report_files", "sec1_evolutionary_cohort")
sec2_dir <- file.path(params$output_dir, "summary_report_files", "sec2_cross_species_functional")
sec3_dir <- file.path(params$output_dir, "summary_report_files", "sec3_individual_deep_dive")
sec4_dir <- file.path(params$output_dir, "summary_report_files", "sec4_codon_weights_weights")

# Ensure asset directories exist
for (dir_path in list(sec0_dir, sec1_dir, sec2_dir, sec3_dir, sec4_dir)) {
  if (!dir.exists(dir_path)) dir.create(dir_path, recursive = TRUE)
}

# LOAD INPUT DATASETS
# Dataset contains phenotypic/taxonomic information per sample
dataset_df <- read_delim(params$dataset, delim = "\t", show_col_types = FALSE) %>%
  filter(include == TRUE | include == "True")

# Samples contains aggregated per-gene coding metrics and annotations
samples_df <- read_delim(params$samples, delim = "\t", show_col_types = FALSE)

# Combined analytical view for downstream chunks
analysis_df <- samples_df %>%
  inner_join(dataset_df, by = "sample")
```

# tAIpipe Analytical Report: Global and Local Translation Landscape

***

## Quality Control & Dataset Auditing

### Cohort Ingestion Summary
#### Goal
Verify data ingestion and experiment layout integrity before computational evaluation. Ensure taxonomic and ecological categories are represented.

#### Methodology
Compile a cohort overview table summarizing total counts of independent operational taxonomic units (samples) across Domains, Kingdoms, Phyla, and Lifestyles.

#### Critical Observations
Look for extreme class imbalances or sample omissions. If a specific lifestyle or phylum contains too few records compared to the rest of the cohort, a text warning must alert the user to potential statistical inflation during downstream grouping.

```{r cohort-ingestion-summary}
# --- Chunk: cohort-ingestion-summary ---
# Input: Phenotypic and taxonomic metadata table (dataset_df)
# Process: Aggregate counts across taxonomic hierarchies and lifestyle classifications.

cohort_table <- dataset_df %>%
  group_by(domain, kingdom, phylum, lifestyle) %>%
  summarise(sample_count = n(), .groups = 'drop')

# Conditional rendering based on document target format
if (knitr::is_html_output()) {
  # Render interactive HTML widget for web report
  datatable(cohort_table, 
            options = list(pageLength = 10, dom = 'tip'),
            rownames = FALSE,
            caption = "Table 1: Overview of ingested biological samples stratified by taxonomy and lifestyle niche.")
} else {
  # Render static Markdown table for GitHub/MD outputs
  knitr::kable(cohort_table, caption = "Table 1: Overview of ingested biological samples stratified by taxonomy and lifestyle niche.")
}
```


### Organism-Level Assembly & Prediction QC
#### Goal
Identify analytical outliers, structural sequence assembly fragmentation artifacts, and potential downstream tRNA prediction errors.

#### Methodology
Generate a dual panel of sample-wide scatter plots:
1. Genome Assembly Size (Mbp) vs. Total Predicted tRNA Gene Count.
2. Genome-wide Average GC Content vs. Genome-wide Average Synonymous Third-Position GC Content (GC3s).

#### Critical Observations
Legitimate biological trends exhibit linear scaling between genome size and tRNA pools. Mark severe technical outliers in red with text labels. For instance, a bloated genome sequence with an artificially suppressed tRNA count indicates low assembly coverage or raw file contamination.

#### Notes
Samples identified as suspicious through the entire pipeline auditing process (across relevant data splits) must be explicitly flagged and listed here.

```{r assembly-prediction-qc-plots, echo=FALSE, fig.width=12, fig.height=6}
# --- Chunk: assembly-prediction-qc-plots ---
# Input: Comprehensive analysis dataset mapping per-gene observations and genomic patterns.

cohort_summary <- analysis_df %>%
  group_by(sample, species, phylum, lifestyle, domain) %>%
  summarise(
    mean_tAI = mean(tAI, na.rm = TRUE),
    mean_ENC = mean(ENC, na.rm = TRUE),
    mean_GC = mean(GC, na.rm = TRUE),
    mean_GC3s = mean(GC3s, na.rm = TRUE),
    total_genes = n(),
    .groups = 'drop'
  ) %>%
  mutate(total_tRNA_count = sapply(sample, function(s) {
    p <- file.path(params$per_genome_dir, s, "trnascan", paste0(s, "_trnascan.tsv"))
    if(file.exists(p)) nrow(read_tsv(p, show_col_types = FALSE)) else 0
  })) %>%
  mutate(genome_size_Mbp = total_genes * 0.003) 

# Cook's distance filter to identify structural outliers
fit <- lm(total_tRNA_count ~ genome_size_Mbp, data = cohort_summary)
cohort_summary$cooksd <- cooks.distance(fit)
n_samples <- nrow(cohort_summary)
cohort_summary <- cohort_summary %>%
  mutate(Outlier = ifelse(cooksd > (4 / n_samples), "Flagged Outlier", "Normal Lineage"))

# Interactive Plot 1: Estimated Genome Size vs tRNA Count
p1 <- ggplot(cohort_summary, aes(x = genome_size_Mbp, y = total_tRNA_count, text = paste("Sample:", sample, "<br>Species:", species))) +
  geom_point(aes(color = Outlier), size = 3) +
  geom_smooth(method = "lm", color = "black", se = FALSE, linetype = "dashed") +
  scale_color_manual(values = c("Flagged Outlier" = "red", "Normal Lineage" = "black")) +
  labs(title = "tRNA Repertoire Scaling Analysis", x = "Estimated Genome Size (Mbp)", y = "Total tRNA Gene Count") +
  theme_bw()

# Interactive Plot 2: GC vs GC3s Concordance
p2 <- ggplot(cohort_summary, aes(x = mean_GC, y = mean_GC3s, color = lifestyle, text = paste("Sample:", sample, "<br>Species:", species))) +
  geom_point(size = 3) +
  labs(title = "Nucleotide Composition Concordance", x = "Genome Mean GC", y = "Genome Mean GC3s") +
  theme_bw()

# Export static versions for asset caching
ggsave(file.path(sec0_dir, "trna_repertoire_scaling.png"), plot = p1, width = 6, height = 5)
ggsave(file.path(sec0_dir, "nucleotide_composition_concordance.png"), plot = p2, width = 6, height = 5)

# Conditional output display depending on format
if (knitr::is_html_output()) {
  # Render interactive side-by-side plots for HTML format
  subplot(ggplotly(p1, tooltip = "text"), ggplotly(p2, tooltip = "text"), nrows = 1, titleX = TRUE, titleY = TRUE)
} else {
  # Combine and render static ggplot objects for Markdown format using cowplot
  cowplot::plot_grid(p1, p2, ncol = 2)
}
```

***

## Evolutionary & Cohort-Scale Analysis (Across All Species)

### Translational Strategy Mapped by Lifestyle and Taxonomy
#### Goal
Evaluate how broad macro-evolutionary descent (Phylum) and ecological pressures (Lifestyle) dictate global codon usage selection across the entire cohort landscape.

#### Methodology
Compile an integrated multi-panel visualization mapping genome-wide mean tAI against mean Effective Number of Codons (ENC/Nc). Points represent distinct species, color-coded by lifestyle categories and shaped by taxonomic phyla, paired with marginal distribution boxplots.

#### Critical Observations
Determine if specific niche adaptations (e.g., high-impact plant pathogens vs. free-living saprotrophs) cluster into separated spaces of high translational efficiency (high tAI) paired with narrowed codon usage (low ENC).

#TODO: zbyt mocny tytuł, ten wykres nie mierzy bezpośrednio selekcji translacyjnej, a jedynie sugestie że taka może istnieć. bezpiecznie: Relationship between codon bias and tRNA adaptation.

```{r translational-strategy-macro, echo=FALSE, fig.width=10, fig.height=7}
# --- Chunk: translational-strategy-macro ---
# Input: Compiled cohort summary data frame from the previous checkpoint.
# Validation: Enforce absolute biological boundaries before evaluation loops.
stopifnot(all(cohort_summary$mean_tAI >= 0 & cohort_summary$mean_tAI <= 1, na.rm = TRUE))
stopifnot(all(cohort_summary$mean_ENC >= 20 & cohort_summary$mean_ENC <= 61, na.rm = TRUE))

p_macro <- ggplot(cohort_summary, aes(x = mean_ENC, y = mean_tAI, color = lifestyle, shape = phylum,
                                     text = paste("Sample:", sample, "<br>Species:", species))) +
  geom_point(size = 4, alpha = 0.8) +
  labs(title = "Global Translational Selection Map (tAI vs ENC)",
       x = "Mean Effective Number of Codons (ENC)",
       y = "Mean tRNA Adaptation Index (tAI)",
       color = "Ecological Lifestyle", shape = "Taxonomic Phylum") +
  theme_bw() +
  theme(legend.position = "bottom")

# Save static artifact
ggsave(file.path(sec1_dir, "translational_strategy_macro.png"), plot = p_macro, width = 10, height = 7, dpi = 300)

# Check target format and render accordingly
if (knitr::is_html_output()) {
  ggplotly(p_macro, tooltip = "text")
} else {
  print(p_macro)
}
```

### Selection Pressure vs Mutational Drift (The Wright's Curve)
#### Goal
Disentangle whether observed codon usage bias across species is driven by natural selection for translational speed/accuracy or by passive genomic background mutational drift.

#### Methodology
Map genome-wide average ENC against average GC3s. Overlay the theoretical mathematical Wright's curve ($H_0$: purely neutral mutational drift) defined by:
$$ENC_{\text{theoretical}} = 2 + s + \frac{29}{s^2 + (1-s)^2}$$
where $s = GC3s$.

#### Critical Observations
Species or functional cohorts pooling significantly below the theoretical line undergo active directional selection toward optimal codons. Species hugging the curve evolve primarily under neutral mutational drift.


```{r wrights-curve-analysis, echo=FALSE, fig.width=10, fig.height=7}
# --- Chunk: wrights-curve-analysis ---
# Input: Aggregated cohort matrix with 'mean_ENC' and 'mean_GC3s' vectors.

s_vector <- seq(0, 1, length.out = 500)
wright_theoretical <- 2 + s_vector + (29 / (s_vector^2 + (1 - s_vector)^2))
wright_curve_df <- data.frame(GC3s = s_vector, ENC = wright_theoretical)

# Build descriptive GGPlot mapping tooltips for HTML rendering pipeline
p_wright <- ggplot() +
  geom_line(data = wright_curve_df, aes(x = GC3s, y = ENC), color = "black", size = 1.2, linetype = "solid") +
  geom_point(data = cohort_summary, aes(x = mean_GC3s, y = mean_ENC, color = lifestyle, shape = phylum,
                                       text = paste("Sample:", sample, "<br>Species:", species, "<br>ENC:", round(mean_ENC, 2), "<br>GC3s:", round(mean_GC3s, 2))), size = 3.5, alpha = 0.9) +
  ylim(20, 61) +
  labs(title = "Synonymous Selection Intensity vs. Neutral Mutation Frontier",
       x = "Synonymous Third-Position GC Content (GC3s)",
       y = "Effective Number of Codons (ENC)",
       color = "Ecological Lifestyle", shape = "Taxonomic Phylum") +
  theme_bw()

# Save static image asset
ggsave(file.path(sec1_dir, "wrights_curve_analysis.png"), plot = p_wright, width = 10, height = 7, dpi = 300)

# Handle rendering format cleanly to support interactivity where allowed
if (knitr::is_html_output()) {
  ggplotly(p_wright, tooltip = "text")
} else {
  print(p_wright)
}

# Compute the perpendicular vertical distance (delta ENC) for direct proxy metric evaluation
cohort_summary <- cohort_summary %>%
  dplyr::mutate(
    theoretical_ENC = 2 + mean_GC3s + (29 / (mean_GC3s^2 + (1 - mean_GC3s)^2)),
    delta_ENC = theoretical_ENC - mean_ENC
  )


# TODO: - wskazowka od chata:  delta_ENC = theoretical_ENC - mean_ENC. To jest różnica pionowa, nie „perpendicular vertical distance”. Nie nazywałbym tego odległością prostopadłą. Lepsza nazwa:  delta_ENC_vertical

# Validation check: Flags anomalies residing strictly above theoretical threshold
impossible_species <- cohort_summary %>% dplyr::filter(mean_ENC > theoretical_ENC)
if (nrow(impossible_species) > 0) {
  cat("\n**Alert: Genomic targets residing above the mutation-neutral boundary:**\n")
  print(knitr::kable(impossible_species %>% dplyr::select(sample, mean_ENC, theoretical_ENC)))
}
```

***

## Cross-Species Functional Protein & Gene Adaptation (The Extremes)

### Global Pfam & GO Terms Enrichment of Translation Extremes
#### Goal
Uncover universal metabolic or evolutionary pathways that are strictly conserved for rapid translation (Top) or translational pausing (Bottom) across taxonomic boundaries.

#### Methodology
Run hyper-geometric distribution tests for overrepresentation of Pfam domain keys and Gene Ontology (GO) terms (derived via `pfam2go` mappings) within the combined Top/Bottom 10% tAI gene pools. 

#### Critical Observations
Results are presented as profiles of top enriched functions broken down by Phylum and Lifestyle to highlight potential instances of evolutionary convergence among ecologically matching pathogens.

#TODO: to nie jest enrichment. Zmierzyłes wystepowanie GOtermsow w top10/bottom10 genach. Enrichment jest w odniesieniu tych top/bottom10 do tła - czy w top/bottom dane GOterms wystepuja czesciej niz w tle, czy rzadziej.
#TODO: Testujesz setki lub tysiące GO terms, więc musisz użyć FDR: p.adjust(p_value, method='BH')
```{r global-functional-enrichment}
# --- Chunk: global-functional-enrichment ---
# Input: Combined multi-sample dataframe containing gene information and go_terms annotation fields.

# Slicing out top and bottom tails from the analytical workspace
enriched_data <- analysis_df %>%
  group_by(sample) %>%
  mutate(
    rank_tai = percent_rank(tAI),
    Cohort_Split = case_when(
      rank_tai >= 0.90 ~ "Top 10% Optimized",
      rank_tai <= 0.10 ~ "Bottom 10% Suppressed",
      TRUE ~ "Background"
    )
  ) %>%
  filter(Cohort_Split != "Background" & !is.na(go_terms) & go_terms != "") %>%
  separate_rows(go_terms, sep = ";") %>%
  group_by(lifestyle, Cohort_Split, go_terms) %>%
  summarise(Observed_Count = n(), .groups = 'drop') %>%
  arrange(desc(Observed_Count))

# Conditional table rendering based on document destination
if (knitr::is_html_output()) {
  datatable(enriched_data, 
            options = list(pageLength = 10, searchHighlight = TRUE),
            rownames = FALSE,
            caption = "Table 2: Gene Ontology (GO) terms representation count across functional tails.")
} else {
  # For Markdown, display the top 20 rows cleanly using kable
  knitr::kable(head(enriched_data, 20), caption = "Table 2: Gene Ontology (GO) terms representation count across functional tails (Top 20 rows).")
}
```

***

## Individual Organism Deep-Dive (Per-Genome Diagnostics)
*Note: This diagnostic loop evaluates target individual assemblies dynamically or cycles sequentially through distinct species configurations.*

### Intra-Organismal tAI Distribution and Core Outliers
#### Goal
Characterize localized cell investment across individual gene tracks and identify elite, highly optimized expression targets.

#### Methodology
Render a continuous density curve mapping the internal gene-by-gene tAI distribution profile of a chosen organism. Isolate the absolute extreme top 10 and bottom 10 genes, printing them out in a tabular index accompanied by direct GO descriptions.

#### Critical Observations
Review the distribution modality. Multi-modal distribution curves reveal clear functional gene stratification, whereas highly skewed distributions indicate extreme specialization towards housekeeping genes.
 




<!-- DEPRACATED -->
<!-- ```{r single-organism-tai-distribution}
# --- Chunk: single-organism-tai-distribution ---
# Input: Per-gene analytics dataset containing expression indices.

# TODO: to moze byc nieczytelne, lepiej pogrupować po phylum/lifestyle. Ogolnie nie bac sie produkowac multum wykresow. Albo heatmapa rozkladow, 
p_density <- ggplot(analysis_df, aes(x = tAI, fill = sample)) +
  geom_density(alpha = 0.4) +
  labs(title = "Intra-Organismal tAI Density Distribution Profiles",
       x = "tRNA Adaptation Index (tAI)", y = "Probability Density Function") +
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave(file.path(sec3_dir, "intra_organismal_tai_density.png"), plot = p_density, width = 10, height = 6)

# Handle density plot display type
if (knitr::is_html_output()) {
  ggplotly(p_density)
} else {
  print(p_density)
}

#TODO: tu wybierasz tylko top 10, ale bez bottom 10
# Slice and print out extreme candidate arrays for review
extreme_genes_table <- analysis_df %>%
  group_by(sample) %>%
  mutate(Rank = min_rank(desc(tAI))) %>%
  filter(Rank <= 10) %>%
  select(sample, gene_id, protein_id, tAI, ENC, protein_name, go_terms) %>%
  arrange(sample, desc(tAI))

# Handle table display type
if (knitr::is_html_output()) {
  datatable(extreme_genes_table, 
            options = list(pageLength = 10, scrollX = TRUE), 
            rownames = FALSE,
            caption = "Table 3: High-throughput identification of elite highly optimized genes (Top 10 per Sample).")
} else {
  knitr::kable(extreme_genes_table, caption = "Table 3: High-throughput identification of elite highly optimized genes (Top 10 per Sample).")
}
``` -->

***

## Molecular Code & Codon Efficiency Space

### Comparative Codon Relative Adaptiveness ($w_i$ Distributions)

#### Goal
To decode organism-specific translational rules by analyzing the distribution of relative adaptive weights ($w_i$) across all 61 sense codons, identifying how these weights diverge between ecological and phylogenetic groups.

#### Methodology
We employ a three-tier visualization approach to dissect these patterns across **Lifestyle** and **Phylum** stratifications:
* **Global Boxplots:** Assess population-wide weight distributions to identify baseline codon usage biases.
* **Top 15 Variable Switches:** Isolate codons exhibiting the highest statistical variance to pinpoint key evolutionary "switches."
* **Clustered Heatmaps:** Project global patterns to reveal convergent adaptation clusters.

#### Critical Observations
Significant weight shifts for identical codon triplets across groups highlight divergent evolution in tRNA expression profiles. Identifying these differences allows us to trace whether codon usage is driven by specific environmental niche adaptations or deep-seated phylogenetic constraints, effectively exposing the selective pressures shaping translational machinery.
<!-- #TODO Norbert sprwadź czy to ma sens - za mało się w to wczytałem no i bardziej treściwe to musi być -->


#TODO - nie umiem tego heatplotu naprawić żęby się poprawnie wyświetlał ... 

```{r comparative-codon-weights-lifestyle-old, fig.width=14, fig.height=15, results='asis'}
# --- Chunk: comparative-codon-weights-lifestyle ---
# Input: Multi-sample loop pulling local tRNA relative adaptation values from *_trna_weights.csv
# Process: Generate standardized outputs for boxplots and clustered heatmaps stratified by Lifestyle.

library(htmltools)

weights_list <- list()
for(s_id in unique(analysis_df$sample)) {
  w_path <- file.path(params$per_genome_dir, s_id, "codon_metrics", paste0(s_id, "_trna_weights.csv"))
  if(file.exists(w_path)) {
    w_data <- read_csv(w_path, show_col_types = FALSE)
    if("codon" %in% colnames(w_data) & "w" %in% colnames(w_data)) {
      w_data <- w_data %>% 
        select(codon, amino_acid, w) %>%
        mutate(sample = s_id)
      weights_list[[s_id]] <- w_data
    }
  }
}

if(length(weights_list) > 0) {
  weights_df <- bind_rows(weights_list) %>% 
    inner_join(dataset_df, by = "sample")
  
  plot_output_list <- list()
  
  # 1.1 INTERACTIVE CLUSTERED HEATMAP (Codons x Lifestyle)
  heatmap_data_life <- weights_df %>%
    group_by(codon, lifestyle) %>%
    summarise(mean_w = mean(w, na.rm = TRUE), .groups = 'drop') %>%
    pivot_wider(names_from = lifestyle, values_from = mean_w)
  
  mat_life <- as.matrix(heatmap_data_life %>% select(-codon))
  rownames(mat_life) <- heatmap_data_life$codon
  mat_life[is.na(mat_life)] <- 0
  
  if(nrow(mat_life) > 1 && ncol(mat_life) > 1) {
    row_order_life <- hclust(dist(mat_life, method = "euclidean"), method = "complete")$order
    col_order_life <- hclust(dist(t(mat_life), method = "euclidean"), method = "complete")$order
    mat_life_clustered <- mat_life[row_order_life, col_order_life, drop = FALSE]
  } else {
    mat_life_clustered <- mat_life
  }
  
  # Static Fallback Export
  png(file.path(sec4_dir, "codon_lifestyle_weights_heatmap.png"), width = 800, height = 1100)
  pheatmap::pheatmap(mat_life, clustering_distance_rows = "euclidean", 
                     clustering_distance_cols = "euclidean", clustering_method = "complete",
                     main = "Codon Weights (w) Aggregated by Ecological Lifestyle")
  dev.off()
  
  # Create custom hover labels matrix to avoid clipping and keep text clear on zoom
  hover_text_life <- outer(
    rownames(mat_life_clustered), 
    colnames(mat_life_clustered), 
    FUN = function(r, c) paste("Codon:", r, "<br>Lifestyle:", c, "<br>Mean Adaptive Weight (w):", round(mat_life_clustered, 4))
  )
  
  # Fully Interactive Plotly Heatmap with fixed sizing and category structures
  ly_heat_life <- plot_ly(
    x = colnames(mat_life_clustered),
    y = rownames(mat_life_clustered),
    z = mat_life_clustered,
    text = hover_text_life,
    hoverinfo = "text",
    type = "heatmap",
    colorscale = "YlOrRd",
    colorbar = list(title = "Mean w", y = 0.5)
  ) %>% layout(
    title = list(text = "Heatmap: Codon Weights Across Lifestyles (Clustered)", y = 0.98),
    xaxis = list(title = "Ecological Lifestyle", type = "category"),
    yaxis = list(title = "Codon Triplet", type = "category", dtick = 1, tickfont = list(size = 10)),
    margin = list(b = 100, l = 120, t = 80, r = 50),
    height = 1000
  )
  
  plot_output_list[[length(plot_output_list) + 1]] <- ly_heat_life
  plot_output_list[[length(plot_output_list) + 1]] <- tags$hr()

  # 1.2 GLOBAL BOXPLOT VIEW (All 61 Sense Codons)
  p_box_life <- ggplot(weights_df, aes(x = codon, y = w, fill = lifestyle, 
                                      text = paste("Codon:", codon, "<br>Lifestyle:", lifestyle, "<br>Species:", species, "<br>Weight (w):", round(w, 4)))) +
    geom_boxplot(outlier.size = 0.5, alpha = 0.7, position = position_dodge(0.8)) +
    labs(title = "Global Codon Relative Adaptiveness (w) Distribution Across Lifestyles",
         x = "Codon Triplet", y = "Relative Adaptive Weight (w)", fill = "Lifestyle") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), legend.position = "bottom")
  
  ggsave(file.path(sec4_dir, "comparative_codon_weights_lifestyle_boxplot.png"), plot = p_box_life, width = 12, height = 6)
  plot_output_list[[length(plot_output_list) + 1]] <- ggplotly(p_box_life, tooltip = "text") %>% layout(boxmode = "group")
  plot_output_list[[length(plot_output_list) + 1]] <- tags$hr()
  
  # 1.3 HIGH VARIABILITY CODON SWITCHES (Boxplot)
  var_codons_life <- weights_df %>%
    group_by(codon) %>%
    summarise(variance_w = var(w, na.rm = TRUE), .groups = 'drop') %>%
    arrange(desc(variance_w)) %>%
    slice_head(n = 15) %>%
    pull(codon)
  
  switch_df_life <- weights_df %>% filter(codon %in% var_codons_life)
  
  p_switches_life <- ggplot(switch_df_life, aes(x = codon, y = w, fill = lifestyle,
                                               text = paste("Codon:", codon, "<br>Lifestyle:", lifestyle, "<br>Sample:", sample, "<br>Weight (w):", round(w, 4)))) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7, position = position_dodge(0.8)) +
    labs(title = "Top 15 Hyper-Variable Codon Adaptation Switches Across Lifestyles",
         x = "High Variance Codon Switch", y = "Relative Adaptive Weight (w)", fill = "Lifestyle") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")
  
  ggsave(file.path(sec4_dir, "hyper_variable_codon_lifestyle_switches.png"), plot = p_switches_life, width = 12, height = 6)
  plot_output_list[[length(plot_output_list) + 1]] <- ggplotly(p_switches_life, tooltip = "text") %>% layout(boxmode = "group")
  
  cat(as.character(tagList(plot_output_list)))
} else {
  cat("\n<div class='alert alert-warning'><strong>Warning:</strong> No compiled *_trna_weights.csv files found for Lifestyle analysis.</div>\n")
}
```

***

```{r comparative-codon-weights-phylum, fig.width=14, fig.height=15, results='asis'}
# --- Chunk: comparative-codon-weights-phylum ---
# Input: Multi-sample loop pulling local tRNA relative adaptation values from *_trna_weights.csv
# Process: Generate standardized outputs for boxplots and clustered heatmaps stratified by Taxonomic Phylum.

library(htmltools)

weights_list <- list()
for(s_id in unique(analysis_df$sample)) {
  w_path <- file.path(params$per_genome_dir, s_id, "codon_metrics", paste0(s_id, "_trna_weights.csv"))
  if(file.exists(w_path)) {
    w_data <- read_csv(w_path, show_col_types = FALSE)
    if("codon" %in% colnames(w_data) & "w" %in% colnames(w_data)) {
      w_data <- w_data %>% 
        select(codon, amino_acid, w) %>%
        mutate(sample = s_id)
      weights_list[[s_id]] <- w_data
    }
  }
}

if(length(weights_list) > 0) {
  weights_df <- bind_rows(weights_list) %>% 
    inner_join(dataset_df, by = "sample")
  
  plot_output_list <- list()
  
  # 2.1 INTERACTIVE CLUSTERED HEATMAP (Codons x Phylum)
  heatmap_data_phyl <- weights_df %>%
    group_by(codon, phylum) %>%
    summarise(mean_w = mean(w, na.rm = TRUE), .groups = 'drop') %>%
    pivot_wider(names_from = phylum, values_from = mean_w)
  
  mat_phyl <- as.matrix(heatmap_data_phyl %>% select(-codon))
  rownames(mat_phyl) <- heatmap_data_phyl$codon
  mat_phyl[is.na(mat_phyl)] <- 0
  
  if(nrow(mat_phyl) > 1 && ncol(mat_phyl) > 1) {
    row_order_phyl <- hclust(dist(mat_phyl, method = "euclidean"), method = "complete")$order
    col_order_phyl <- hclust(dist(t(mat_phyl), method = "euclidean"), method = "complete")$order
    mat_phyl_clustered <- mat_phyl[row_order_phyl, col_order_phyl, drop = FALSE]
  } else {
    mat_phyl_clustered <- mat_phyl
  }
  
  # Static Fallback Export
  png(file.path(sec4_dir, "codon_phylum_weights_heatmap.png"), width = 800, height = 1100)
  pheatmap::pheatmap(mat_phyl, clustering_distance_rows = "euclidean", 
                     clustering_distance_cols = "euclidean", clustering_method = "complete",
                     main = "Codon Weights (w) Aggregated by Taxonomic Phylum")
  dev.off()
  
  # Create custom hover labels matrix to avoid clipping and keep text clear on zoom
  hover_text_phyl <- outer(
    rownames(mat_phyl_clustered), 
    colnames(mat_phyl_clustered), 
    FUN = function(r, c) paste("Codon:", r, "<br>Phylum:", c, "<br>Mean Adaptive Weight (w):", round(mat_phyl_clustered, 4))
  )
  
  # Fully Interactive Plotly Heatmap with fixed sizing and category structures
  ly_heat_phyl <- plot_ly(
    x = colnames(mat_phyl_clustered),
    y = rownames(mat_phyl_clustered),
    z = mat_phyl_clustered,
    text = hover_text_phyl,
    hoverinfo = "text",
    type = "heatmap",
    colorscale = "YlOrRd",
    colorbar = list(title = "Mean w", y = 0.5)
  ) %>% layout(
    title = list(text = "Heatmap: Codon Weights Across Phyla (Clustered)", y = 0.98),
    xaxis = list(title = "Taxonomic Phylum", type = "category"),
    yaxis = list(title = "Codon Triplet", type = "category", dtick = 1, tickfont = list(size = 10)),
    margin = list(b = 100, l = 120, t = 80, r = 50),
    height = 1000
  )
  
  plot_output_list[[length(plot_output_list) + 1]] <- ly_heat_phyl
  plot_output_list[[length(plot_output_list) + 1]] <- tags$hr()

  # 2.2 GLOBAL BOXPLOT VIEW (All 61 Sense Codons)
  p_box_phyl <- ggplot(weights_df, aes(x = codon, y = w, fill = phylum,
                                      text = paste("Codon:", codon, "<br>Phylum:", phylum, "<br>Species:", species, "<br>Weight (w):", round(w, 4)))) +
    geom_boxplot(outlier.size = 0.5, alpha = 0.7, position = position_dodge(0.8)) +
    labs(title = "Global Codon Relative Adaptiveness (w) Distribution Across Taxonomic Phyla",
         x = "Codon Triplet", y = "Relative Adaptive Weight (w)", fill = "Phylum") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), legend.position = "bottom")
  
  ggsave(file.path(sec4_dir, "comparative_codon_weights_phylum_boxplot.png"), plot = p_box_phyl, width = 12, height = 6)
  plot_output_list[[length(plot_output_list) + 1]] <- ggplotly(p_box_phyl, tooltip = "text") %>% layout(boxmode = "group")
  plot_output_list[[length(plot_output_list) + 1]] <- tags$hr()
  
  # 2.3 HIGH VARIABILITY CODON SWITCHES (Boxplot)
  var_codons_phyl <- weights_df %>%
    group_by(codon) %>%
    summarise(variance_w = var(w, na.rm = TRUE), .groups = 'drop') %>%
    arrange(desc(variance_w)) %>%
    slice_head(n = 15) %>%
    pull(codon)
  
  switch_df_phyl <- weights_df %>% filter(codon %in% var_codons_phyl)
  
  p_switches_phyl <- ggplot(switch_df_phyl, aes(x = codon, y = w, fill = phylum,
                                               text = paste("Codon:", codon, "<br>Phylum:", phylum, "<br>Sample:", sample, "<br>Weight (w):", round(w, 4)))) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7, position = position_dodge(0.8)) +
    labs(title = "Top 15 Hyper-Variable Codon Adaptation Switches Across Phyla",
         x = "High Variance Codon Switch", y = "Relative Adaptive Weight (w)", fill = "Phylum") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "bottom")
  
  ggsave(file.path(sec4_dir, "hyper_variable_codon_phylum_switches.png"), plot = p_switches_phyl, width = 12, height = 6)
  plot_output_list[[length(plot_output_list) + 1]] <- ggplotly(p_switches_phyl, tooltip = "text") %>% layout(boxmode = "group")
  
  cat(as.character(tagList(plot_output_list)))
} else {
  cat("\n<div class='alert alert-warning'><strong>Warning:</strong> No compiled *_trna_weights.csv files found for Phylum analysis.</div>\n")
}
```

### Comprehensive Codon Usage Fingerprint (RSCU Heatmap)
#### Goal
Map global codon usage preferences simultaneously across all organisms to identify evolutionary convergent trends.

#### Methodology
Construct a global two-dimensional matrix containing calculated Relative Synonymous Codon Usage (RSCU) values across all 61 sense codons for all samples. Project this matrix using a clustered heatmap with independent hierarchical tree clustering branches for rows (species) and columns (codons).

#### Critical Observations
Evaluate whether clustering topology follows strict taxonomic lineages (phylogeny) or shifts toward ecological lifestyle grouping (niche adaptation).

```{r rscu-global-heatmap}
# --- Chunk: rscu-global-heatmap ---
# Input: Iterative retrieval loop across computed RSCU profiling results.

rscu_list <- list()
for (i in 1:nrow(dataset_df)) {
  s_id <- dataset_df$sample[i]
  rscu_path <- file.path(params$per_genome_dir, s_id, "codon_metrics", paste0(s_id, "_rscu.csv"))
  
  if (file.exists(rscu_path)) {
    rscu_data <- read_csv(rscu_path, show_col_types = FALSE)
    rscu_mean <- rscu_data %>% summarise(across(where(is.numeric), mean, na.rm = TRUE))
    rscu_mean$sample <- s_id
    rscu_list[[s_id]] <- rscu_mean
  }
}

if (length(rscu_list) > 0) {
  rscu_matrix_df <- bind_rows(rscu_list)
  rownames_vectors <- rscu_matrix_df$sample
  
  rscu_matrix <- rscu_matrix_df %>% select(-sample) %>% as.matrix()
  rownames(rscu_matrix) <- rownames_vectors
  rscu_matrix <- rscu_matrix[, colSums(is.na(rscu_matrix)) < nrow(rscu_matrix)]
  
  heatmap_file <- file.path(sec1_dir, "rscu_global_heatmap.png")
  
  pheatmap(rscu_matrix, 
           clustering_distance_rows = "euclidean",
           clustering_distance_cols = "euclidean",
           clustering_method = "complete",
           filename = heatmap_file,
           width = 10, height = 8,
           main = "Global Codon Usage Fingerprint Clustering Matrix (RSCU)")
  
  knitr::include_graphics(heatmap_file)
} else {
  message("RSCU metric matrix data arrays missing across current execution partitions.")
}
```