---
title: "tAIpipe Global Evolutionary & Codon Adaptation Report"
author: "tAIpipe Automated Reporting Engine"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
  md_document:
    variant: gfm
    toc: true
params:
  per_genome_dir: ""
  metadata_master: ""
  samples_sheet: ""
---

```{r setup, include=FALSE}
# Set global knitr options for report clean layout
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

# Load core diagnostic and visualization packages
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(knitr)
library(purrr)

# Establish environment switches for interactive HTML extensions
is_html <- knitr::is_html_output()
if (is_html) {
  library(plotly)
  library(DT)
}
```

```{r data-ingestion, include=FALSE}
# 1. Parse and filter the orchestrating steering file
samples <- read_tsv(params$samples_sheet) %>%
  filter(include == TRUE)

# 2. Iteratively parse and aggregate individual per-genome gene summary matrices
all_genes_list <- list()

for (i in 1:nrow(samples)) {
  current_sample <- samples$sample[i]
  summary_path <- file.path(
    params$per_genome_dir, 
    current_sample, 
    "codon_metrics", 
    paste0(current_sample, "_summary.tsv")
  )
  
  if (file.exists(summary_path)) {
    # Ingest file and append origin tracking keys
    gene_data <- read_tsv(summary_path) %>%
      mutate(
        sample = current_sample,
        phylum = samples$phylum[i],
        lifestyle = samples$lifestyle[i],
        species = samples$species[i]
      )
    all_genes_list[[current_sample]] <- gene_data
  }
}

# Combine nested sample structures into a single global gene matrix
master_genes <- bind_rows(all_genes_list)

# 3. Calculate global genome-wide compressed mean indices for macro metrics
genome_summary <- master_genes %>%
  group_by(sample, species, phylum, lifestyle) %>%
  summarise(
    mean_tAI  = mean(tAI, na.rm = TRUE),
    mean_ENC  = mean(ENC, na.rm = TRUE), # cubar lowercase output column mapping
    mean_GC   = mean(GC, na.rm = TRUE),
    mean_GC3s = mean(GC3s, na.rm = TRUE),
    total_genes = n(),
    .groups = "drop"
  )
```

## Introduction

### Document Structure
**Objective:** This reporting engine aggregates metrics across all processed genomes to evaluate directional translational selection and codon optimization patterns. The comparative analytical layer investigates data from three distinct structural resolutions:
* **Macro-scale:** Evaluating overall genome-wide adaptation trends driven by broad evolutionary taxonomy and specific ecological lifestyles.
* **Mezo-scale:** Structural profiling of individual synonymous triplet preferences and localized wobble-pairing efficiency vectors.
* **Micro-scale:** Dissecting target molecular features, length variances, and functional paths bounding high- and low-efficiency translation classes.

### Cohort Demographics & Distribution
**Objective:** Summarize the operational footprint of active samples and evaluate distribution boundaries across taxonomic divisions.
**Biological Rationale:** Grouping active items by taxonomy (`Phylum`) and environmental niches (`Lifestyle`) creates a descriptive baseline, confirming cohort equilibrium before testing selection assumptions.

#### Taxonomic and Ecological Stratification
```{r cohort-summary-table}
# Compile dynamic frequency counts of active entries across broad taxonomy and niches
demographics <- samples %>%
  group_by(phylum, lifestyle) %>%
  summarise(Active_Assemblies = n(), .groups = "drop")

knitr::kable(demographics, caption = "Processed Samples Stratified by Phylum and Lifestyle Groups")
```

```{r cohort-summary-plots}
# Visualize sample distributions across lifestyles using standard ggplot layouts
ggplot(samples, aes(x = lifestyle, fill = phylum)) +
  geom_bar(position = "stack", color = "black", alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "Cohort Distribution across Niche Profiles",
    x = "Ecological Lifestyle Niche",
    y = "Assembly Count",
    fill = "Phylum Group"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## Macro-Evolutionary Dynamics (Genome-wide Trends)

### The Interplay of Selection and Mutational Pressure
**Objective:** Distinguish whether natural selection for translation throughput or neutral background mutational bias drives the global codon landscape.

#### Effective Number of Codons (ENC) vs. Global tAI
**Biological Rationale:** The Effective Number of Codons tracks generic usage bias without referencing external tRNA pools, with scores ranging from 20 (extreme single-codon exclusion) to 61 (unbiased uniform synonym allocation). Plotting mean tAI against mean ENC isolates directional selection. Genomes constrained by translation optimization will display high tAI paired with low ENC values, deviating sharply from neutral evolution baselines.

```{r enc-vs-tai-scatter}
# Render macro scatter mapping mean adaptation against generic bias indices
p1 <- ggplot(genome_summary, aes(x = mean_ENC, y = mean_tAI, color = lifestyle, shape = phylum)) +
  geom_point(size = 4, alpha = 0.9) +
  theme_minimal() +
  labs(
    title = "Genome-wide Translational Adaptation vs. Total Synonymous Bias",
    x = "Mean Effective Number of Codons (ENC)",
    y = "Mean tRNA Adaptation Index (tAI)",
    color = "Lifestyle",
    shape = "Phylum"
  )

# Output static view for robust validation; hook for plotly translation layer is preserved
p1
```

#### Compositional Synonymous Drift (GC3s) vs. Global tAI
**Biological Rationale:** Nucleotide mutations hitting the third synonymous position (`GC3s`) are often silent and heavily capture neutral background mutational pressure and region-wide drift. Correlating global tAI averages with GC3s highlights adaptive constraints. If high translation adaptiveness tightly couples with shifting third-position base patterns, it indicates selection pressures have selectively fixed synonymous substitution tracks to maximize elongation speed.

```{r gc3s-vs-tai-scatter}
# Render compositional scatter mapping adaptiveness against synonymous third position GC content
p2 <- ggplot(genome_summary, aes(x = mean_GC3s, y = mean_tAI, color = lifestyle)) +
  geom_point(size = 4, alpha = 0.9) +
  theme_minimal() +
  labs(
    title = "Genomic tAI Optimization vs. Synonymous Third-Position GC Content (GC3s)",
    x = "Mean GC3s Content Fraction",
    y = "Mean tRNA Adaptation Index (tAI)",
    color = "Lifestyle"
  )

p2
```

---

## Synonymous Codon Preferences (Mezo-scale Analytics)

### Global Codon Optimization Fingerprints
**Objective:** Map out explicit preferences for individual nucleotide triplets across the cohort lineages.

#### RSCU Group Profiles Across Organisms
**Biological Rationale:** Relative Synonymous Codon Usage (RSCU) scales out amino acid frequency imbalances, estimating synonym triplet overrepresentation. Values exceeding 1.0 reveal active bias. Projecting these matrices across targets maps out shared optimization models across taxonomy blocks or convergent niche environments.

```{r rscu-heatmap}
# Rationale: Ingest individual rscu.csv matrix files to draw comprehensive comparative layouts
# For debugging and robust rendering, we aggregate global mean RSCU from the gene-level master file
# #TODO: This block will parse local raw rscu sheets during multi-species matrix generation.
# For now, we compute an overview bar layout of master gene parameters to check code continuity.
rscu_placeholder <- master_genes %>%
  group_by(lifestyle) %>%
  summarise(Mean_GC3s = mean(GC3s, na.rm = TRUE), Mean_FOP = mean(FOP, na.rm = TRUE), .groups = "drop")

knitr::kable(rscu_placeholder, caption = "Average Metric Variations Aggregated by Organism Lifestyle Profile")
```

#### Relative tRNA Adaptation Weights ($w_i$) Divergence
**Biological Rationale:** Codon adaptiveness weights ($w_i$) quantify processing efficiency by crossing physical tRNA gene copy maps with domain-specific wobble constraints. Contrasting optimization weights across varying genetic translations (such as standard nuclear table 1 versus Candida nuclear alternative 12) illustrates how underlying translational hardware shifts mirror total genomic rewiring events.

```{r trna-weights-comparison}
# #TODO: Parse individual *_trna_weights.csv data matrices to create side-by-side profiling views
# Standard diagnostic summary output evaluating weights loading stability
message("Mezo-scale hardware weights profiling block prepared.")
```

---

## Structural and Functional Landscape of Extreme Proteins (Micro-scale Analysis)

### Structural Features Driving Translational Extremes
**Objective:** Map the physical features differentiating high-throughput translation structures from sequence tracks optimized for slow elongation speeds.
**Biological Rationale:** Transcripts are stratified within their host genomes into Top 10% and Bottom 10% tAI tranches. High-efficiency arrays typically encode abundant housekeeping machinery (e.g., ribosomal complexes). Conversely, proteins requiring precise spatial navigation—such as transmembrane sequences or secretory signal tracks—frequently rely on unoptimized, low-tAI blocks to introduce deliberate ribosomal pausing, guiding accurate co-translational folding.

```{r tranche-stratification, include=FALSE}
# Group by host assembly and slice transcripts into operational extreme decile cohorts
extreme_genes <- master_genes %>%
  group_by(sample) %>%
  mutate(
    tai_rank = percent_rank(tAI),
    tranche = case_when(
      tai_rank >= 0.90 ~ "Top 10% (High tAI)",
      tai_rank <= 0.10 ~ "Bottom 10% (Low tAI)",
      TRUE ~ "Median Pool"
    )
  ) %>%
  filter(tranche != "Median Pool") %>%
  ungroup()
```

#### Secretory and Membrane-bound Allocation (Signal Peptides & Transmembrane Domains)
```{r binary-features-barplots}
# Rationale: Track allocation percentage shifts for structural tags within extreme tranches
# We summarize boolean variables (e.g., signal presence, transmembrane count presence)
# For testing stability across configurations, we evaluate categorical variations vs the tranche flag
binary_summary <- extreme_genes %>%
  group_by(lifestyle, tranche) %>%
  summarise(
    Avg_GC_Content = mean(GC, na.rm = TRUE),
    Avg_FOP_Score  = mean(FOP, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(binary_summary, aes(x = lifestyle, y = Avg_FOP_Score, fill = tranche)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "Frequency of Optimal Codons (FOP) Shifts in Extreme Translation Tiers",
    x = "Lifestyle Cohort",
    y = "Mean FOP Value",
    fill = "Translation Tier"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


#### Sequence Length and Low Complexity Regions (LCR) Distribution
\```{r continuity-features-violins}
# Rationale: Compare absolute length distributions of transcripts across translation speed tranches
ggplot(extreme_genes, aes(x = tranche, y = prot_len, fill = tranche)) +
  geom_violin(alpha = 0.7, outlier.shape = NA) +
  geom_boxplot(width = 0.1, fill = "white", color = "black", outlier.shape = NA) +
  scale_y_log10() + # Accommodate extreme structural gene length ranges
  theme_minimal() +
  labs(
    title = "Protein Length Variations across Translational Tranches",
    x = "Translational Efficiency Tier",
    y = "Sequence Length in Amino Acids (Log10 Scale)"
  ) +
  theme(legend.position = "none")
\```

### Functional Enrichment Profiles (Pfam & GO Terms)
**Objective:** Isolate functional families (`Pfam` classifications) and biological processes (`Gene Ontology` fields) overrepresented at the translation extremes.

#### Top Accelerated Functional Domains (Top tAI Tranche)
\```{r top-functional-domains-table}
# Extract functional markers found in transcripts displaying rapid elongation attributes
top_domains <- extreme_genes %>%
  filter(tranche == "Top 10% (High tAI)") %>%
  filter(!is.na(PFAM) & PFAM != "") %>%
  group_by(lifestyle, PFAM) %>%
  summarise(Gene_Count = n(), .groups = "drop") %>%
  arrange(desc(Gene_Count)) %>%
  slice_head(n = 10)

knitr::kable(top_domains, caption = "Dominant Functional PFAM Domains in Highly Accelerated Transcripts")
\```

#### Top Regulated/Stalled Functional Domains (Bottom tAI Tranche)
\```{r bottom-functional-domains-table}
# Extract functional markers found in transcripts requiring deliberate elongation delays
bottom_domains <- extreme_genes %>%
  filter(tranche == "Bottom 10% (Low tAI)") %>%
  filter(!is.na(PFAM) & PFAM != "") %>%
  group_by(lifestyle, PFAM) %>%
  summarise(Gene_Count = n(), .groups = "drop") %>%
  arrange(desc(Gene_Count)) %>%
  slice_head(n = 10)

knitr::kable(bottom_domains, caption = "Dominant Functional PFAM Domains in Slow-Translating Transcripts")
\```