---
title: "tAIpipe Comprehensive Analytical Report"
subtitle: "Global and Local Translation Optimization Landscape"
author: "tAIpipe Automated Framework"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: flatly
    highlight: tango
  md_document:
    variant: markdown_github
params:
  samples: ""
  dataset: ""
  per_genome_dir: ""
  output_dir: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 7)

library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
library(cowplot)
library(pheatmap)

# STANDARDIZED SECTIONS PATH DEFINITIONS
# Define and create output subdirectories for asset isolation
sec0_dir <- file.path(params$output_dir, "summary_report_files", "sec0_quality_control")
sec1_dir <- file.path(params$output_dir, "summary_report_files", "sec1_evolutionary_cohort")
sec2_dir <- file.path(params$output_dir, "summary_report_files", "sec2_cross_species_functional")
sec3_dir <- file.path(params$output_dir, "summary_report_files", "sec3_individual_deep_dive")
sec4_dir <- file.path(params$output_dir, "summary_report_files", "sec4_molecular_code")

dir.create(sec0_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(sec1_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(sec2_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(sec3_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(sec4_dir, recursive = TRUE, showWarnings = FALSE)
```

# tAIpipe Analytical Report: Global and Local Translation Landscape

---

## Quality Control & Dataset Auditing

### Cohort Ingestion Summary
#### Goal
Verify data ingestion and experiment layout integrity before computational evaluation. Ensure taxonomic and ecological categories are represented.

#### Methodology
Compile a cohort overview table summarizing total counts of independent operational taxonomic units (samples) across Domains, Kingdoms, Phyla, and Lifestyles.

#### Critical Observations
Look for extreme class imbalances or sample omissions. If a specific lifestyle or phylum contains too few records compared to the rest of the cohort, a text warning must alert the user to potential statistical inflation during downstream grouping.

```{r cohort-ingestion-summary, echo=FALSE, results='asis'}
# --- Chunk: cohort-ingestion-summary ---
# Input: Read the primary multi-sample manifest 'samples_test.tsv' and phenotypic table 'test_dataset.tsv'.
# Process: Group rows by 'domain', 'kingdom', 'phylum', and 'lifestyle'. Compute total active samples (include == True).

# Read and isolate active samples from the design sheet configuration parameter
samples_df <- readr::read_tsv(params$dataset, show_col_types = FALSE) %>%
  dplyr::filter(include == TRUE | include == "True")

summary_table <- samples_df %>%
  dplyr::group_by(domain, kingdom, phylum, lifestyle) %>%
  dplyr::summarise(Total_Samples = dplyr::n(), .groups = 'drop')

# Return: A summary markdown table using knitr::kable().
print(knitr::kable(summary_table, caption = "Summary of Successfully Ingested Cohort Groups"))

# Validation: Trigger a programmatic alert/warning text if any active group size drops below N=3 (statistical risk).
low_representation_groups <- summary_table %>% dplyr::filter(Total_Samples < 3)

if (nrow(low_representation_groups) > 0) {
  cat("\n<div class='alert alert-warning'><strong>⚠️ WARNING:</strong> The following sample cohorts have fewer than 3 active biological representatives. High risk of statistical inflation during macro comparative groupings:</div>\n")
  print(knitr::kable(low_representation_groups))
}
```

### Organism-Level Assembly & Prediction QC
#### Goal
Identify analytical outliers, structural sequence assembly fragmentation artifacts, and potential downstream tRNA prediction errors.

#### Methodology
Generate a dual panel of sample-wide scatter plots:
1. Genome Assembly Size (Mbp) vs. Total Predicted tRNA Gene Count.
2. Genome-wide Average GC Content vs. Genome-wide Average Synonymous Third-Position GC Content (GC3s).

#### Critical Observations
Legitimate biological trends exhibit linear scaling between genome size and tRNA pools. Mark severe technical outliers in red with text labels. For instance, a bloated genome sequence with an artificially suppressed tRNA count indicates low assembly coverage or raw file contamination.

#### Notes
Samples identified as suspicious through the entire pipeline auditing process (across relevant data splits) must be explicitly flagged and listed here.

```{r assembly-prediction-qc-plots, echo=FALSE, fig.width=12, fig.height=6}
# --- Chunk: assembly-prediction-qc-plots ---
# Input: Global cohort metadata table ('cohort_summary.tsv') containing aggregated single-genome indices.
master_df <- readr::read_tsv(params$samples, show_col_types = FALSE)

# Generate single-genome summary attributes dynamically from the master table
cohort_summary <- master_df %>%
  dplyr::group_by(sample, phylum, lifestyle) %>%
  dplyr::summarise(
    mean_tAI = mean(tAI, na.rm = TRUE),
    mean_ENC = mean(ENC, na.rm = TRUE),
    mean_GC = mean(GC, na.rm = TRUE),
    mean_GC3s = mean(GC3s, na.rm = TRUE),
    total_genes = dplyr::n(),
    .groups = 'drop'
  ) %>%
  dplyr::mutate(total_tRNA_count = safe_trna_count <- sapply(sample, function(s) {
    p <- file.path(params$per_genome_dir, s, "trnascan", paste0(s, "_trnascan.tsv"))
    if(file.exists(p)) nrow(readr::read_tsv(p, show_col_types = FALSE)) else 0
  })) %>%
  # Approximate physical size in Mbp based on total annotated open reading frame sets
  dplyr::mutate(genome_size_Mbp = total_genes * 0.003) 

# Process: Identify points with Cook's distance > 4/n.
fit <- lm(total_tRNA_count ~ genome_size_Mbp, data = cohort_summary)
cohort_summary$cooksd <- cooks.distance(fit)
n_samples <- nrow(cohort_summary)
cohort_summary <- cohort_summary %>%
  dplyr::mutate(Outlier = ifelse(cooksd > (4 / n_samples), "Flagged Outlier", "Normal Lineage"))

# Plot 1: X = genome_size_Mbp, Y = total_tRNA_count. Fit a linear regression model.
p1 <- ggplot(cohort_summary, aes(x = genome_size_Mbp, y = total_tRNA_count)) +
  geom_point(aes(color = Outlier), size = 3) +
  geom_smooth(method = "lm", color = "black", se = FALSE, linetype = "dashed") +
  geom_text(data = filter(cohort_summary, Outlier == "Flagged Outlier"), aes(label = sample), vjust = -1, color = "red") +
  scale_color_manual(values = c("Flagged Outlier" = "red", "Normal Lineage" = "black")) +
  labs(title = "tRNA Repertoire Scaling", x = "Estimated Genome Size (Mbp)", y = "Total tRNA Gene Count") +
  theme_bw()

# Plot 2: X = mean_GC, Y = mean_GC3s.
p2 <- ggplot(cohort_summary, aes(x = mean_GC, y = mean_GC3s, color = lifestyle)) +
  geom_point(size = 3) +
  labs(title = "Nucleotide Composition Concordance", x = "Genome Mean GC", y = "Genome Mean GC3s") +
  theme_bw()

# Return: Render a side-by-side graphical display (e.g., using cowplot or patchwork).
qc_panel <- cowplot::plot_grid(p1, p2, ncol = 2)
print(qc_panel)

# Export structural static graphics targets


outlier_list <- cohort_summary %>% dplyr::filter(Outlier == "Flagged Outlier") %>% dplyr::pull(sample)
if (length(outlier_list) > 0) {
  cat("\n**Suspicious Assemblies Flagged for Anomalous tRNA/Size Ratio:**\n")
  cat(paste("-", outlier_list, collapse = "\n"), "\n")
}
```

---

## Evolutionary & Cohort-Scale Analysis (Across All Species)

### Translational Strategy Mapped by Lifestyle and Taxonomy
#### Goal
Evaluate how broad macro-evolutionary descent (Phylum) and ecological pressures (Lifestyle) dictate global codon usage selection across the entire cohort landscape.

#### Methodology
Compile an integrated multi-panel visualization mapping genome-wide mean tAI against mean Effective Number of Codons (ENC/Nc). Points represent distinct species, color-coded by lifestyle categories and shaped by taxonomic phyla, paired with marginal distribution boxplots.

#### Critical Observations
Determine if specific niche adaptations (e.g., high-impact plant pathogens vs. free-living saprotrophs) cluster into separated spaces of high translational efficiency (high tAI) paired with narrowed codon usage (low ENC).

```{r translational-strategy-macro, echo=FALSE, fig.width=10, fig.height=7}
# --- Chunk: translational-strategy-macro ---
# Input: Combined cohort metric file containing mean values for every validated genome.
# Process: Create a scatter plot mapping X = mean_ENC, Y = mean_tAI. Add marginal boxplots for 'lifestyle' and 'phylum'.

# Validation: Assert that no parsed metric exceeds biological boundaries (tAI between 0 and 1; ENC between 20 and 61).
stopifnot(all(cohort_summary$mean_tAI >= 0 & cohort_summary$mean_tAI <= 1))
stopifnot(all(cohort_summary$mean_ENC >= 20 & cohort_summary$mean_ENC <= 61))

p_macro <- ggplot(cohort_summary, aes(x = mean_ENC, y = mean_tAI, color = lifestyle, shape = phylum)) +
  geom_point(size = 4, alpha = 0.8) +
  labs(title = "Global Translational Selection Map (tAI vs ENC)",
       x = "Mean Effective Number of Codons (ENC)",
       y = "Mean tRNA Adaptation Index (tAI)",
       color = "Ecological Lifestyle", shape = "Taxonomic Phylum") +
  theme_bw() +
  theme(legend.position = "bottom")

# Return: An integrated scatter-marginal-boxplot assembly.
print(p_macro)

# Save high resolution image file
ggsave(file.path(sec1_dir, "translational_strategy_macro.png"), plot = p_macro, width = 10, height = 7, dpi = 300)
```

### Selection Pressure vs Mutational Drift (The Wright's Curve)
#### Goal
Disentangle whether observed codon usage bias across species is driven by natural selection for translational speed/accuracy or by passive genomic background mutational drift.

#### Methodology
Map genome-wide average ENC against average GC3s. Overlay the theoretical mathematical Wright's curve ($H_0$: purely neutral mutational drift) defined by:
$$ENC_{\text{theoretical}} = 2 + s + \frac{29}{s^2 + (1-s)^2}$$
where $s = GC3s$.

#### Critical Observations
Species or functional cohorts pooling significantly below the theoretical line undergo active directional selection toward optimal codons. Species hugging the curve evolve primarily under neutral mutational drift.

```{r wrights-curve-analysis, echo=FALSE, fig.width=10, fig.height=7}
# --- Chunk: wrights-curve-analysis ---
# Input: Aggregated cohort matrix with 'mean_ENC' and 'mean_GC3s' vectors.
# Process: Generate a continuous math function vector tracking Wright's equation across s in [0, 1].
s_vector <- seq(0, 1, length.out = 500)
wright_theoretical <- 2 + s_vector + (29 / (s_vector^2 + (1 - s_vector)^2))
wright_curve_df <- data.frame(GC3s = s_vector, ENC = wright_theoretical)

# Plot actual species data points over the theoretical curve line.
p_wright <- ggplot() +
  geom_line(data = wright_curve_df, aes(x = GC3s, y = ENC), color = "black", size = 1.2, linetype = "solid") +
  geom_point(data = cohort_summary, aes(x = mean_GC3s, y = mean_ENC, color = lifestyle, shape = phylum), size = 3.5, alpha = 0.9) +
  ylim(20, 61) +
  labs(title = "Synonymous Selection Intensity vs. Neutral Mutation Frontier",
       x = "Synonymous Third-Position GC Content (GC3s)",
       y = "Effective Number of Codons (ENC)") +
  theme_bw()

print(p_wright)
ggsave(file.path(sec1_dir, "wrights_curve_analysis.png"), plot = p_wright, width = 10, height = 7, dpi = 300)

# Compute the perpendicular vertical distance (delta ENC) for each species as a direct proxy for selection intensity.
cohort_summary <- cohort_summary %>%
  dplyr::mutate(
    theoretical_ENC = 2 + mean_GC3s + (29 / (mean_GC3s^2 + (1 - mean_GC3s)^2)),
    delta_ENC = theoretical_ENC - mean_ENC
  )

# Validation: Flags and records any species located above the mathematical boundary line.
impossible_species <- cohort_summary %>% dplyr::filter(mean_ENC > theoretical_ENC)
if (nrow(impossible_species) > 0) {
  cat("\n**Alert: Genomic targets residing above the mutation-neutral boundary:**\n")
  print(knitr::kable(impossible_species %>% dplyr::select(sample, mean_ENC, theoretical_ENC)))
}
```

---

## Type-Level Functional Protein & Gene Adaptation (The Extremes)

### Structural Features of Protein Extremes Across Cohorts
#### Goal
Determine global amino acid and protein structural sorting rules relative to translational efficiency across the entire experimental framework.

#### Methodology
Aggregate gene sequences from all genomes into a pooled dataset. Isolate top 10% (optimized) and bottom 10% (slowed/regulatory) global tAI thresholds. Construct dodge bar charts displaying the fraction of genes containing signal peptides or transmembrane domains (TM), grouped by Phylum or Lifestyle.

#### Spatial Profiling
For the Top 10% and Bottom 10% tAI groups, generate density profile charts across 10 unified length bins tracking the positional layout of:
* Predicted protein domain boundaries (Pfam)
* Transmembrane segments (TMHMM)
* Low Complexity Regions (LCR)
This directly checks if highly optimized proteins restrict sequence complexity or transmembrane blocks to specific structural domains (e.g., N- or C-termini) compared to slower translational targets.

```{r cross-species-structural-extremes, echo=FALSE, fig.width=12, fig.height=10}
# --- Chunk: cross-species-structural-extremes ---
# Input: Pooled structural metric file containing gene-by-gene measurements across all organisms ('master_gene_table.tsv').
global_thresholds <- quantile(master_df$tAI, probs = c(0.1, 0.9), na.rm = TRUE)

# Filter and slice global Top 10% and Bottom 10% sub-cohorts.
master_df <- master_df %>%
  dplyr::mutate(tAI_Group = case_when(
    tAI <= global_thresholds[1] ~ "Bottom 10% (Slow)",
    tAI >= global_thresholds[2] ~ "Top 10% (Fast)",
    TRUE ~ "Background"
  ))

extremes_df <- master_df %>% dplyr::filter(tAI_Group %in% c("Bottom 10% (Slow)", "Top 10% (Fast)"))

# Compute percentage presence for Signal Peptides and TM domains across groups.
bar_data <- extremes_df %>%
  dplyr::group_by(tAI_Group, phylum) %>%
  dplyr::summarise(
    Signal_Peptide = mean(`Presence of signal peptide; Boolean` == TRUE | `Presence of signal peptide; Boolean` == "True", na.rm = TRUE) * 100,
    Transmembrane = mean(`number of transmembrane elements predicted with TMHMM` > 0, na.rm = TRUE) * 100,
    .groups = 'drop'
  ) %>%
  tidyr::pivot_longer(cols = c(Signal_Peptide, Transmembrane), names_to = "Feature", values_to = "Percentage")

p_bars <- ggplot(bar_data, aes(x = phylum, y = Percentage, fill = tAI_Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~Feature) +
  labs(title = "Structural Feature Slicing Across Global tAI Boundaries", x = "Phylum", y = "Proportion of Genes (%)", fill = "Translational Space") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Validation: Confirm that input sequence lengths are properly normalized (0.0 to 1.0) before bin assignment.
extract_spatial_bins <- function(df, column_name, label) {
  split_lists <- strsplit(as.character(df[[column_name]]), ",")
  matrix_bins <- do.call(rbind, lapply(split_lists, function(x) {
    vals <- as.numeric(x)
    if(length(vals) != 10) rep(0, 10) else vals
  }))
  colnames(matrix_bins) <- paste0("Bin", 1:10)
  
  cbind(df %>% dplyr::select(tAI_Group), matrix_bins) %>%
    tidyr::pivot_longer(cols = starts_with("Bin"), names_to = "Bin", values_to = "Density") %>%
    dplyr::mutate(Bin = as.numeric(sub("Bin", "", Bin)), Feature = label)
}

lcr_spatial <- extract_spatial_bins(extremes_df, "Symbolic localization of LCR; 10 bins scaled to sum up to total protein length", "Low Complexity Regions (LCR)")
tm_spatial <- extract_spatial_bins(extremes_df, "Symbolic localization of transmembrane elements; 10 bins scaled to sum up to total protein length", "Transmembrane Domains (TM)")

combined_spatial <- rbind(lcr_spatial, tm_spatial) %>%
  dplyr::group_by(tAI_Group, Bin, Feature) %>%
  dplyr::summarise(Mean_Density = mean(Density, na.rm = TRUE), .groups = 'drop')

p_spatial <- ggplot(combined_spatial, aes(x = Bin, y = Mean_Density, color = tAI_Group)) +
  geom_line(size = 1.2) +
  facet_wrap(~Feature, scales = "free_y") +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "Spatial Metagene Feature Profiling Along Normalized Protein Length Coordinates", x = "Structural Position Bin (N-term to C-term)", y = "Mean Density Accumulation Score") +
  theme_bw()

composite_structural <- cowplot::plot_grid(p_bars, p_spatial, ncol = 1, rel_heights = c(1, 1))
print(composite_structural)

ggsave(file.path(sec2_dir, "cross_species_structural_extremes.png"), plot = p_bars, width = 10, height = 5, dpi = 300)
ggsave(file.path(sec2_dir, "cross_species_spatial_profiles.png"), plot = p_spatial, width = 12, height = 5, dpi = 300)
```

### Global Pfam & GO Terms Enrichment of Translation Extremes
#### Goal
Uncover universal metabolic or evolutionary pathways that are strictly conserved for rapid translation (Top) or translational pausing (Bottom) across taxonomic boundaries.

#### Methodology
Run hyper-geometric distribution tests for overrepresentation of Pfam domain keys and Gene Ontology (GO) terms (derived via `pfam2go` mappings) within the combined Top/Bottom 10% tAI gene pools. 

#### Critical Observations
Results are presented as profiles of top enriched functions broken down by Phylum and Lifestyle to highlight potential instances of evolutionary convergence among ecologically matching pathogens.

```{r global-functional-enrichment, echo=FALSE, results='asis'}
# --- Chunk: global-functional-enrichment ---
# Input: Global sliced gene indices linked with mapped Pfam IDs and GO terms.
go_parsing_df <- master_df %>%
  dplyr::select(tAI_Group, phylum, `GO terms based on Pfam domains obtained by mapping on pfam2go, separated with the pipe symbol '|'`) %>%
  dplyr::rename(GO_raw = 3) %>%
  dplyr::filter(!is.na(GO_raw) & GO_raw != "") %>%
  tidyr::separate_rows(GO_raw, sep = "\\|")

# Execute a hyper-geometric enrichment routine setup
universe_counts <- go_parsing_df %>% dplyr::group_by(GO_raw) %>% dplyr::summarise(Universe_N = dplyr::n())
group_counts <- go_parsing_df %>% dplyr::group_by(tAI_Group, phylum, GO_raw) %>% dplyr::summarise(Count_Selected = dplyr::n(), .groups = 'drop')

# Confirm that the background universe set matches total valid protein-coding sequences ingested by the workflow.
total_universe_n <- nrow(go_parsing_df)

enrichment_output <- group_counts %>%
  dplyr::filter(tAI_Group %in% c("Bottom 10% (Slow)", "Top 10% (Fast)")) %>%
  dplyr::left_join(universe_counts, by = "GO_raw") %>%
  dplyr::mutate(
    Sample_Size = ifelse(tAI_Group == "Top 10% (Fast)", sum(master_df$tAI_Group == "Top 10% (Fast)"), sum(master_df$tAI_Group == "Bottom 10% (Slow)")),
    p_value = phyper(Count_Selected - 1, Universe_N, total_universe_n - Universe_N, Sample_Size, lower.tail = FALSE)
  ) %>%
  dplyr::mutate(adj_p_value = p.adjust(p_value, method = "BH")) %>%
  dplyr::filter(adj_p_value < 0.05) %>%
  dplyr::arrange(adj_p_value)

print(knitr::kable(head(enrichment_output, 20), caption = "Top 20 Broad Functional GO Terms Enriched in Translation Rate Extremes"))

# FIXED: Redirect CSV export away from workflow/scripts/ straight into output directories
write_csv(enrichment_output, file.path(sec2_dir, "global_functional_enrichment_results.csv"))
```

---

## Individual Organism Deep-Dive (Per-Genome Diagnostics) {.tabset}
*Note: This diagnostic loop evaluates target individual assemblies dynamically or cycles sequentially through distinct species configurations.*

```{r deep-dive-config, include=FALSE}
# Isolate first sample key from master data to automate evaluation dynamically without manual breaks
target_sample_id <- unique(master_df$sample)[1]
sample_isolated_df <- master_df %>% dplyr::filter(sample == target_sample_id)
```


```{r local-organism-deep-dive-loop, echo=FALSE, results='asis', fig.width=14, fig.height=10}
# --- Chunk: local-organism-deep-dive-loop ---
# Input: Read master table using the updated parameter name integrated from Snakemake
master_df <- readr::read_tsv(params$samples, show_col_types = FALSE)

# Resolve target reports directory dynamically relative to per_genome_dir path from config
output_reports_dir <- file.path(dirname(params$per_genome_dir), "aggregated", "reports")
dir.create(output_reports_dir, showWarnings = FALSE, recursive = TRUE)

unique_samples <- unique(master_df$sample)

for (target_sample_id in unique_samples) {
  
  # Isolate specific data split for the current organism iteration
  local_df <- master_df %>%
    dplyr::filter(sample == target_sample_id) %>%
    dplyr::rename(
      Prot_Len = `Protein length`,
      TM_Count = `number of transmembrane elements predicted with TMHMM`,
      TM_Len = `total length of transmembrane elements`,
      LCR_Count = `Total number of LCR`,
      LCR_Len = `Total length of LCR`
    )
  
  if (nrow(local_df) == 0) next
  
  # Compute local boundary slices dynamically for the current genome distribution
  local_bounds <- quantile(local_df$tAI, probs = c(0.1, 0.9), na.rm = TRUE)
  
  local_df <- local_df %>%
    dplyr::mutate(Local_Cohort = dplyr::case_when(
      tAI <= local_bounds[1] ~ "Local Bottom 10% (Slow)",
      tAI >= local_bounds[2] ~ "Local Top 10% (Fast)",
      TRUE ~ "Core Genome Background"
    ))
  
  plot_subset <- local_df %>% 
    dplyr::filter(Local_Cohort %in% c("Local Bottom 10% (Slow)", "Local Top 10% (Fast)"))
  
  # Create a clean tab set container per organism for the HTML dashboard interface
  cat("\n### ", target_sample_id, " \n")
  
  cat("\n#### Intra-Organismal tAI Distribution and Core Outliers\n")
  
  # Generate continuous density line plot object
  p_local_density <- ggplot(local_df, aes(x = tAI)) +
    geom_density(fill = "aquamarine", alpha = 0.4, color = "darkgreen", size = 1.2) +
    labs(title = paste("Intra-Organismal Capacity Allocation Profile:", target_sample_id), 
         x = "Gene tAI Score Metric", y = "Kernel Density") +
    theme_bw()
  
  print(p_local_density)
  cat("\n\n")
  
  # Enforce programmatic saving directly to the correct section folder location
  ggsave(file.path(sec3_dir, paste0(target_sample_id, "_tai_local_density.png")), 
         plot = p_local_density, width = 10, height = 5, dpi = 300)
  
  # Sort table to harvest expression extremes and resolve NA/empty annotation outputs safely
  top_10_genes <- local_df %>% dplyr::arrange(desc(tAI)) %>% head(10)
  bottom_10_genes <- local_df %>% dplyr::arrange(tAI) %>% head(10)
  
  go_col_name <- colnames(local_df)[grepl("GO terms", colnames(local_df))]
  
  clean_print_table <- function(sub_df) {
    sub_df %>% 
      dplyr::select(`Protein ID (NCBI accession)`, tAI, ENC, dplyr::all_of(go_col_name)) %>%
      dplyr::rename(GO_Terms = 4) %>%
      dplyr::mutate(GO_Terms = dplyr::case_when(
        is.na(GO_Terms) | GO_Terms == "" | GO_Terms == "NA" ~ "Unannotated / No Pfam Domain Found",
        TRUE ~ GO_Terms
      ))
  }
  
  cat("\n##### Top 10 High-Efficiency Expression Target Genes (Elite Pool)\n")
  print(knitr::kable(clean_print_table(top_10_genes), format = "markdown"))
  cat("\n\n")
  
  cat("\n##### Bottom 10 Regulatory/Slowing Translation Genes (Structural Control)\n")
  print(knitr::kable(clean_print_table(bottom_10_genes), format = "markdown"))
  cat("\n\n")
  
  cat("\n#### Local Protein Feature Correlation\n")
  
  # Structural violin and box plots definitions
  p_len <- ggplot(plot_subset, aes(x = Local_Cohort, y = Prot_Len, fill = Local_Cohort)) +
    geom_violin(alpha = 0.6, trim = TRUE) +
    geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
    labs(title = "Protein Length Space", x = "", y = "Length (Amino Acids)") +
    theme_bw() + theme(legend.position = "none")
  
  p_lcr_counts <- ggplot(plot_subset, aes(x = Local_Cohort, y = LCR_Count, fill = Local_Cohort)) +
    geom_boxplot(alpha = 0.7, outlier.colour = "red", outlier.size = 1) +
    labs(title = "LCR Density Bounds", x = "", y = "Number of Detected LCRs") +
    theme_bw() + theme(legend.position = "none")
  
  p_tm_len <- ggplot(plot_subset, aes(x = Local_Cohort, y = TM_Len, fill = Local_Cohort)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    labs(title = "Total Transmembrane Span Length", x = "", y = "Total TM Length (AA)") +
    theme_bw() + theme(legend.position = "none")
  
  # Spatial tracking transformations for 3 sub-bins arrays mapping
  lcr_3bin_col <- local_df %>% dplyr::select(dplyr::contains("3 bins")) %>% colnames()
  lcr_3bin_strings <- plot_subset %>% dplyr::select(Local_Cohort, dplyr::all_of(lcr_3bin_col)) %>% dplyr::rename(B3 = 2)
  
  split_3bins <- strsplit(as.character(lcr_3bin_strings$B3), ",")
  matrix_3bins <- do.call(rbind, lapply(split_3bins, function(x) {
    v <- as.numeric(x)
    if(length(v) != 3 || any(is.na(v))) rep(0, 3) else v
  }))
  colnames(matrix_3bins) <- c("N-terminus (0-0.25)", "Middle (0.25-0.75)", "C-terminus (0.75-1.0)")
  
  polarization_df <- cbind(lcr_3bin_strings %>% dplyr::select(Local_Cohort), matrix_3bins) %>%
    tidyr::pivot_longer(cols = -Local_Cohort, names_to = "Region", values_to = "LCR_Accumulation") %>%
    dplyr::group_by(Local_Cohort, Region) %>%
    dplyr::summarise(Mean_Accumulation = mean(LCR_Accumulation, na.rm = TRUE), .groups = 'drop')
  
  p_polar_bars <- ggplot(polarization_df, aes(x = Region, y = Mean_Accumulation, fill = Local_Cohort)) +
    geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
    labs(title = "Spatial LCR Polarization Profile", x = "Functional Structural Region", y = "Mean Residue Length") +
    theme_bw() + theme(legend.position = "bottom")
  
  local_panel <- cowplot::plot_grid(
    cowplot::plot_grid(p_len, p_lcr_counts, p_tm_len, ncol = 3),
    p_polar_bars,
    ncol = 1,
    rel_heights = c(1, 1.2)
  )
  print(local_panel)
  cat("\n\n")
  
  # Redirect generated panel graphic straight to the correct section folder
  ggsave(file.path(sec3_dir, paste0(target_sample_id, "_local_structural_panel.png")), 
         plot = local_panel, width = 14, height = 10, dpi = 300)
  
  cat("\n#### Overlap of Structural Domains and Sequence Complexity\n")
  overlap_column_name <- master_df %>% dplyr::select(dplyr::contains("overlapping with LCRs")) %>% colnames()
  overlap_matrix <- local_df %>%
    dplyr::filter(Local_Cohort %in% c("Local Bottom 10% (Slow)", "Local Top 10% (Fast)")) %>%
    dplyr::select(`Protein ID (NCBI accession)`, Local_Cohort, dplyr::all_of(overlap_column_name)) %>%
    dplyr::rename(Overlapping_Domains = 3) %>%
    dplyr::mutate(Overlapping_Domains = dplyr::case_when(
      is.na(Overlapping_Domains) | Overlapping_Domains == "" ~ "No Overlap Detected (>80%)",
      TRUE ~ Overlapping_Domains
    ))
  
  print(knitr::kable(head(overlap_matrix, 20), format = "markdown"))
  cat("\n\n")
}
```

### Intra-Organismal tAI Distribution and Core Outliers
#### Goal
Characterize localized cell investment across individual gene tracks and identify elite, highly optimized expression targets.

#### Methodology
Render a continuous density curve mapping the internal gene-by-gene tAI distribution profile of a chosen organism. Isolate the absolute extreme top 10 and bottom 10 genes, printing them out in a tabular index accompanied by direct GO descriptions.

#### Critical Observations
Review the distribution modality. Multi-modal distribution curves reveal clear functional gene stratification, whereas highly skewed distributions indicate extreme specialization towards housekeeping genes.

```{r single-organism-tai-distribution, echo=FALSE, results='asis', fig.width=10, fig.height=5}
# --- Chunk: single-organism-tai-distribution ---
# Input: The standalone sample evaluation track summary ('results/per_genome/<sample>/codon_metrics/*_summary.tsv').

# Generate an empirical density distribution plot for local gene tAI values.
p_local_density <- ggplot(sample_isolated_df, aes(x = tAI)) +
  geom_density(fill = "aquamarine", alpha = 0.4, color = "darkgreen", size = 1.2) +
  labs(title = paste("Intra-Organismal Core Translation Profile Capacity Allocation:", target_sample_id), x = "Gene tAI Value", y = "Density Distribution Parameter") +
  theme_bw()

print(p_local_density)
ggsave(file.path(sec3_dir, paste0(target_sample_id, "_tai_local_density.png")), plot = p_local_density, width = 10, height = 5, dpi = 300)

# Sort the table ascending/descending to harvest the highest 10 and lowest 10 tAI gene IDs.
top_10_genes <- sample_isolated_df %>% dplyr::arrange(desc(tAI)) %>% head(10)
bottom_10_genes <- sample_isolated_df %>% dplyr::arrange(tAI) %>% head(10)

# Confirm that top and bottom slices contain unique gene IDs and check for empty GO annotation strings.
cat("\n#### Top 10 High-Efficiency Expression Target Genes (Elite Pool)\n")
print(knitr::kable(top_10_genes %>% dplyr::select(`Protein ID (NCBI accession)`, tAI, ENC, `GO terms based on Pfam domains obtained by mapping on pfam2go, separated with the pipe symbol '|'`), format = "markdown"))

cat("\n#### Bottom 10 Regulatory/Slowing Translation Genes (Structural Control/Folding Delays)\n")
print(knitr::kable(bottom_10_genes %>% dplyr::select(`Protein ID (NCBI accession)`, tAI, ENC, `GO terms based on Pfam domains obtained by mapping on pfam2go, separated with the pipe symbol '|'`), format = "markdown"))
```

### Local Protein Feature Correlation
#### Goal
Examine localized structural design principles inside a single organism to check if translational speed optimization matches structural folding requirements and LCR management.

#### Methodology
Build violin/boxplot charts evaluating protein length, total TM domain count, total TM domain length, LCR counts, and absolute LCR lengths against Top 10% vs. Bottom 10% subsets.

#### N/C-Terminal LCR Polarization
Construct bar charts comparing LCR density and specific amino acid compositions across three distinct functional regions: N-terminus (0.0-0.25), Internal (0.25-0.75), and C-terminus (0.75-1.0). This verifies whether low-complexity sequences localized at the N-terminus correlate with reduced tAI to implement a controlled ribosomal translation initiation delay.

```{r local-protein-feature-correlation, echo=FALSE, fig.width=14, fig.height=10}
# --- Chunk: local-protein-feature-correlation ---
# Input: Selected organism summary track integrated with structural locus annotations.
# Safe Column Mapping to avoid string breaks and handle NA values gracefully
local_df <- master_df %>%
  dplyr::filter(sample == target_sample_id) %>%
  dplyr::rename(
    Prot_Len = `Protein length`,
    TM_Count = `number of transmembrane elements predicted with TMHMM`,
    TM_Len = `total length of transmembrane elements`,
    LCR_Count = `Total number of LCR`,
    LCR_Len = `Total length of LCR`
  )

local_bounds <- quantile(local_df$tAI, probs = c(0.1, 0.9), na.rm = TRUE)

local_df <- local_df %>%
  dplyr::mutate(Local_Cohort = dplyr::case_when(
    tAI <= local_bounds[1] ~ "Local Bottom 10% (Slow)",
    tAI >= local_bounds[2] ~ "Local Top 10% (Fast)",
    TRUE ~ "Core Genome Background"
  ))

plot_subset <- local_df %>% 
  dplyr::filter(Local_Cohort %in% c("Local Bottom 10% (Slow)", "Local Top 10% (Fast)"))

# Plot 1: Violin chart for Protein Length
p_len <- ggplot(plot_subset, aes(x = Local_Cohort, y = Prot_Len, fill = Local_Cohort)) +
  geom_violin(alpha = 0.6, trim = TRUE) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  labs(title = "Protein Length Distribution", x = "", y = "Length (Amino Acids)") +
  theme_bw() + 
  theme(legend.position = "none")

# [NEW PLOT 1]: Distribution of Total LCR Counts per Gene across local cohorts
p_lcr_counts <- ggplot(plot_subset, aes(x = Local_Cohort, y = LCR_Count, fill = Local_Cohort)) +
  geom_boxplot(alpha = 0.7, outlier.colour = "red", outlier.size = 1) +
  labs(title = "LCR Abundance per Protein", x = "", y = "Number of Detected LCRs") +
  theme_bw() + 
  theme(legend.position = "none")

# [NEW PLOT 2]: Transmembrane Domain Length Boxplot
p_tm_len <- ggplot(plot_subset, aes(x = Local_Cohort, y = TM_Len, fill = Local_Cohort)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  labs(title = "Total Transmembrane Span Length", x = "", y = "Total TM Length (AA)") +
  theme_bw() + 
  theme(legend.position = "none")

# Process 3-Bin Spatial Polarization for LCRs
# FIXED: Safe column extraction using standard base R execution layer
selected_cols <- local_df %>% dplyr::select(Local_Cohort, dplyr::contains("3 bins"))
lcr_3bin_col <- colnames(selected_cols)[-1]

lcr_3bin_strings <- plot_subset %>%
  dplyr::select(Local_Cohort, dplyr::all_of(lcr_3bin_col)) %>%
  dplyr::rename(B3 = 2)

split_3bins <- strsplit(as.character(lcr_3bin_strings$B3), ",")
matrix_3bins <- do.call(rbind, lapply(split_3bins, function(x) {
  v <- as.numeric(x)
  if(length(v) != 3 || any(is.na(v))) rep(0, 3) else v
}))
colnames(matrix_3bins) <- c("N-terminus (0-0.25)", "Middle (0.25-0.75)", "C-terminus (0.75-1.0)")

polarization_df <- cbind(lcr_3bin_strings %>% dplyr::select(Local_Cohort), matrix_3bins) %>%
  tidyr::pivot_longer(cols = -Local_Cohort, names_to = "Region", values_to = "LCR_Accumulation") %>%
  dplyr::group_by(Local_Cohort, Region) %>%
  dplyr::summarise(Mean_Accumulation = mean(LCR_Accumulation, na.rm = TRUE), .groups = 'drop')

p_polar_bars <- ggplot(polarization_df, aes(x = Region, y = Mean_Accumulation, fill = Local_Cohort)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  labs(title = "Spatial LCR Polarization Mapping", x = "Functional Structural Domain Region", y = "Mean Residue Length in Bins") +
  theme_bw() +
  theme(legend.position = "bottom")

# Return: Panel displaying violin charts alongside localized structural bar plots.
local_structural_panel <- cowplot::plot_grid(
  cowplot::plot_grid(p_len, p_lcr_counts, p_tm_len, ncol = 3),
  p_polar_bars,
  ncol = 1,
  rel_heights = c(1, 1.2)
)
print(local_structural_panel)

# Enforce clean handling of zero values and save static PNG files
ggsave(file.path(sec3_dir, paste0(target_sample_id, "_local_structural_panel.png")), plot = local_structural_panel, width = 14, height = 10, dpi = 300)
```

### Overlap of Structural Domains and Sequence Complexity
#### Goal
Determine if low complexity sequences (LCRs) directly collide with functional structural folds within high-efficiency translation frameworks.

#### Methodology
Calculate the physical intersection length between predicted Pfam domain bounds and localized LCR coordinates. Flag structural elements experiencing an intersection overlap of greater than 80% total length across extreme tAI brackets.

#### Critical Observations
High overlap scores in high-tAI genes point to structural optimization constraints, whereas high overlap in low-tAI genes could signify regulatory regions.

```{r domain-lcr-structural-overlap, echo=FALSE, results='asis'}
# --- Chunk: domain-lcr-structural-overlap ---
# Input: Base coordinate track files mapping Pfam intervals and LCR intervals for the target genome.
# Process: Compute interval overlap lengths: Filter rows matching overlap >= 80%.
overlap_column_name <- master_df %>% dplyr::select(dplyr::contains("overlapping with LCRs")) %>% colnames()

overlap_matrix <- local_df %>%
  dplyr::filter(Local_Cohort %in% c("Local Bottom 10% (Slow)", "Local Top 10% (Fast)")) %>%
  dplyr::select(`Protein ID (NCBI accession)`, Local_Cohort, dplyr::all_of(overlap_column_name)) %>%
  dplyr::rename(Overlapping_Domains = 3) %>%
  # Handle empty or NA strings dynamically to ensure tidy table generation
  dplyr::mutate(Overlapping_Domains = dplyr::case_when(
    is.na(Overlapping_Domains) | Overlapping_Domains == "" | Overlapping_Domains == "Not Available" ~ "No Overlap Detected (>80%)",
    TRUE ~ Overlapping_Domains
  ))

# Return: Matrix table displaying highly intersected functional domains grouped by extreme tAI status.
# Validation: Verify that coordinate index tracking systems are unified across input tracks.
print(knitr::kable(head(overlap_matrix, 20), caption = paste("Identified Severe Domain-LCR Intersections for", target_sample_id)))
```

---

## Molecular Code & Codon Efficiency Space

### Comparative Codon Relative Adaptiveness ($w_i$ Distributions)
#### Goal
Decode the underlying organismal translation rules by assessing how specific codon adaptation weights vary across lifestyles or engineering groups.

#### Methodology
Compile distribution charts of calculated relative adaptiveness weights ($w_i$) for all 61 sense codons. Contrast these profiles across ecological classifications or domains.

#### Critical Observations
Identify shifting optimal codon switches. If distinct biological groups display contrasting weights for identical codons, it highlights divergent evolution in their underlying tRNA expression profiles.

```{r comparative-codon-weights, echo=FALSE, fig.width=14, fig.height=8}
# --- Chunk: comparative-codon-weights ---
# Input: Processed individual codon weight tables ('*_trna_weights.csv') compiled across genomes.
weight_files_list <- list.files(path = params$per_genome_dir, pattern = "_trna_weights.csv", recursive = TRUE, full.names = TRUE)

compiled_weights_list <- lapply(weight_files_list, function(filepath) {
  sample_key <- basename(dirname(dirname(filepath)))
  df <- readr::read_csv(filepath, show_col_types = FALSE)
  if (ncol(df) >= 2) {
    colnames(df)[1:2] <- c("Codon", "Weight")
    df$sample <- sample_key
    return(df)
  }
  return(NULL)
})

dataset_meta <- readr::read_tsv(params$dataset, show_col_types = FALSE) %>%
  dplyr::select(sample, lifestyle, phylum)

weights_master_df <- do.call(rbind, Filter(Negate(is.null), compiled_weights_list)) %>%
  dplyr::left_join(dataset_meta, by = "sample")

p_weights_dist <- ggplot(weights_master_df, aes(x = Codon, y = Weight, fill = lifestyle)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  labs(title = "Codon Relative Adaptiveness Weights (w_i) Partitioned Across Lifestyles", 
       x = "Sense Codon Triplets", y = "Relative Adaptation Weight (w_i)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 9, face = "bold"),
        legend.position = "bottom")

print(p_weights_dist)
ggsave(file.path(sec4_dir, "global_codon_weights_distribution.png"), plot = p_weights_dist, width = 14, height = 8, dpi = 300)
```

### Comprehensive Codon Usage Fingerprint (RSCU Heatmap)
#### Goal
Map global codon usage preferences simultaneously across all organisms to identify evolutionary convergent trends.

#### Methodology
Construct a global two-dimensional matrix containing calculated Relative Synonymous Codon Usage (RSCU) values across all 61 sense codons for all samples. Project this matrix using a clustered heatmap with independent hierarchical tree clustering branches for rows (species) and columns (codons).

#### Critical Observations
Evaluate whether clustering topology follows strict taxonomic lineages (phylogeny) or shifts toward ecological lifestyle grouping (niche adaptation).

```{r rscu-global-heatmap, echo=FALSE, fig.width=12, fig.height=9}
# --- Chunk: rscu-global-heatmap ---
# Input: Global RSCU profiles matrices tracks extraction.
rscu_files_list <- list.files(path = params$per_genome_dir, pattern = "_rscu.csv", recursive = TRUE, full.names = TRUE)

compiled_rscu_list <- lapply(rscu_files_list, function(filepath) {
  sample_key <- basename(dirname(dirname(filepath)))
  df <- readr::read_csv(filepath, show_col_types = FALSE)
  if (ncol(df) >= 2) {
    colnames(df)[1:2] <- c("Codon", "RSCU")
    df$sample = sample_key
    return(df)
  }
  return(NULL)
})

rscu_master_df <- do.call(rbind, Filter(Negate(is.null), compiled_rscu_list))

# Execute a hierarchical clustering routing with forced numeric casting and NA cleanup
heatmap_matrix <- rscu_master_df %>%
  dplyr::mutate(Rscu_Numeric = as.numeric(RSCU)) %>%
  dplyr::mutate(Rscu_Numeric = ifelse(is.na(Rscu_Numeric) | is.nan(Rscu_Numeric), 0.0, Rscu_Numeric)) %>%
  tidyr::pivot_wider(id_cols = sample, names_from = Codon, values_from = Rscu_Numeric, values_fill = 0.0) %>%
  as.data.frame()

rownames(heatmap_matrix) <- heatmap_matrix$sample
heatmap_matrix <- as.matrix(heatmap_matrix[, -1])

# Filter out non-informative columns with zero variance across active samples
variance_vector <- apply(heatmap_matrix, 2, sd)
heatmap_matrix_filtered <- heatmap_matrix[, variance_vector > 0, drop = FALSE]

if (ncol(heatmap_matrix_filtered) == 0) {
  heatmap_matrix_filtered <- heatmap_matrix
}

# Dynamic and robust break generation execution layer
mat_min <- min(heatmap_matrix_filtered, na.rm = TRUE)
mat_max <- max(heatmap_matrix_filtered, na.rm = TRUE)

if (abs(mat_max - mat_min) < 1e-5) {
  mat_min <- mat_min - 0.1
  mat_max <- mat_max + 0.1
}

custom_breaks <- seq(mat_min, mat_max, length.out = 51)
dynamic_palette <- colorRampPalette(c("navy", "white", "firebrick3"))(50)

# FIXED: Save matrix heatmap directly inside the target section output path
png(file.path(sec4_dir, "rscu_global_hierarchical_heatmap.png"), width = 1200, height = 900, res = 130)
pheatmap::pheatmap(heatmap_matrix_filtered, 
                    clustering_distance_rows = "euclidean", 
                    clustering_distance_cols = "euclidean", 
                    clustering_method = "complete",
                    scale = "none",
                    color = dynamic_palette,
                    breaks = custom_breaks,
                    legend = TRUE,
                    main = "Global Codon Usage Fingerprint (Hierarchical Clustered RSCU Profiles)")
dev.off()

pheatmap::pheatmap(heatmap_matrix_filtered, 
                    clustering_distance_rows = "euclidean", 
                    clustering_distance_cols = "euclidean", 
                    clustering_method = "complete",
                    scale = "none",
                    color = dynamic_palette,
                    breaks = custom_breaks,
                    legend = TRUE,
                    main = "Global Codon Usage Fingerprint (Hierarchical Clustered RSCU Profiles)")
```