---
title: "tAIpipe Comprehensive Analytical Report"
subtitle: "Global and Local Translation Optimization Landscape"
author: "tAIpipe Automated Framework"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: flatly
    highlight: tango
  md_document:
    variant: markdown_github
params:
  samples: ""
  dataset: ""
  per_genome_dir: ""
  output_dir: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 7)

library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
library(cowplot)
library(pheatmap)

# STANDARDIZED SECTIONS PATH DEFINITIONS
# Define and create output subdirectories for asset isolation
sec0_dir <- file.path(params$output_dir, "summary_report_files", "sec0_quality_control")
sec1_dir <- file.path(params$output_dir, "summary_report_files", "sec1_evolutionary_cohort")
sec2_dir <- file.path(params$output_dir, "summary_report_files", "sec2_cross_species_functional")
sec3_dir <- file.path(params$output_dir, "summary_report_files", "sec3_individual_deep_dive")
sec4_dir <- file.path(params$output_dir, "summary_report_files", "sec4_molecular_code")

dir.create(sec0_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(sec1_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(sec2_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(sec3_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(sec4_dir, recursive = TRUE, showWarnings = FALSE)
```

# tAIpipe Analytical Report: Global and Local Translation Landscape

---

## Quality Control & Dataset Auditing

### Cohort Ingestion Summary
#### Goal
Verify data ingestion and experiment layout integrity before computational evaluation. Ensure taxonomic and ecological categories are represented.

#### Methodology
Compile a cohort overview table summarizing total counts of independent operational taxonomic units (samples) across Domains, Kingdoms, Phyla, and Lifestyles.

#### Critical Observations
Look for extreme class imbalances or sample omissions. If a specific lifestyle or phylum contains too few records compared to the rest of the cohort, a text warning must alert the user to potential statistical inflation during downstream grouping.



```{r assembly-prediction-qc-plots, echo=FALSE, fig.width=12, fig.height=6}
# --- Chunk: assembly-prediction-qc-plots ---
# Input: Global cohort metadata table ('cohort_summary.tsv') containing aggregated single-genome indices.
samples_df <- readr::read_tsv(params$samples, show_col_types = FALSE)

# Generate single-genome summary attributes dynamically from the master table
cohort_summary <- samples_df %>%
  dplyr::group_by(sample, phylum, lifestyle) %>%
  dplyr::summarise(
    mean_tAI = mean(tAI, na.rm = TRUE),
    mean_ENC = mean(ENC, na.rm = TRUE),
    mean_GC = mean(GC, na.rm = TRUE),
    mean_GC3s = mean(GC3s, na.rm = TRUE),
    total_genes = dplyr::n(),
    .groups = 'drop'
  ) %>%
  dplyr::mutate(total_tRNA_count = safe_trna_count <- sapply(sample, function(s) {
    p <- file.path(params$per_genome_dir, s, "trnascan", paste0(s, "_trnascan.tsv"))
    if(file.exists(p)) nrow(readr::read_tsv(p, show_col_types = FALSE)) else 0
  })) %>%
  # Approximate physical size in Mbp based on total annotated open reading frame sets
  dplyr::mutate(genome_size_Mbp = total_genes * 0.003) 

# Process: Identify points with Cook's distance > 4/n.
fit <- lm(total_tRNA_count ~ genome_size_Mbp, data = cohort_summary)
cohort_summary$cooksd <- cooks.distance(fit)
n_samples <- nrow(cohort_summary)
cohort_summary <- cohort_summary %>%
  dplyr::mutate(Outlier = ifelse(cooksd > (4 / n_samples), "Flagged Outlier", "Normal Lineage"))

# Plot 1: X = genome_size_Mbp, Y = total_tRNA_count. Fit a linear regression model.
p1 <- ggplot(cohort_summary, aes(x = genome_size_Mbp, y = total_tRNA_count)) +
  geom_point(aes(color = Outlier), size = 3) +
  geom_smooth(method = "lm", color = "black", se = FALSE, linetype = "dashed") +
  geom_text(data = filter(cohort_summary, Outlier == "Flagged Outlier"), aes(label = sample), vjust = -1, color = "red") +
  scale_color_manual(values = c("Flagged Outlier" = "red", "Normal Lineage" = "black")) +
  labs(title = "tRNA Repertoire Scaling", x = "Estimated Genome Size (Mbp)", y = "Total tRNA Gene Count") +
  theme_bw()

# Plot 2: X = mean_GC, Y = mean_GC3s.
p2 <- ggplot(cohort_summary, aes(x = mean_GC, y = mean_GC3s, color = lifestyle)) +
  geom_point(size = 3) +
  labs(title = "Nucleotide Composition Concordance", x = "Genome Mean GC", y = "Genome Mean GC3s") +
  theme_bw()

# Return: Render a side-by-side graphical display (e.g., using cowplot or patchwork).
qc_panel <- cowplot::plot_grid(p1, p2, ncol = 2)
print(qc_panel)

# Export structural static graphics targets


outlier_list <- cohort_summary %>% dplyr::filter(Outlier == "Flagged Outlier") %>% dplyr::pull(sample)
if (length(outlier_list) > 0) {
  cat("\n**Suspicious Assemblies Flagged for Anomalous tRNA/Size Ratio:**\n")
  cat(paste("-", outlier_list, collapse = "\n"), "\n")
}
```

### Organism-Level Assembly & Prediction QC
#### Goal
Identify analytical outliers, structural sequence assembly fragmentation artifacts, and potential downstream tRNA prediction errors.

#### Methodology
Generate a dual panel of sample-wide scatter plots:
1. Genome Assembly Size (Mbp) vs. Total Predicted tRNA Gene Count.
2. Genome-wide Average GC Content vs. Genome-wide Average Synonymous Third-Position GC Content (GC3s).

#### Critical Observations
Legitimate biological trends exhibit linear scaling between genome size and tRNA pools. Mark severe technical outliers in red with text labels. For instance, a bloated genome sequence with an artificially suppressed tRNA count indicates low assembly coverage or raw file contamination.

#### Notes
Samples identified as suspicious through the entire pipeline auditing process (across relevant data splits) must be explicitly flagged and listed here.

```{r translational-strategy-macro, echo=FALSE, fig.width=10, fig.height=7}
# --- Chunk: translational-strategy-macro ---
# Input: Combined cohort metric file containing mean values for every validated genome.
# Process: Create a scatter plot mapping X = mean_ENC, Y = mean_tAI. Add marginal boxplots for 'lifestyle' and 'phylum'.

# Validation: Assert that no parsed metric exceeds biological boundaries (tAI between 0 and 1; ENC between 20 and 61).
stopifnot(all(cohort_summary$mean_tAI >= 0 & cohort_summary$mean_tAI <= 1))
stopifnot(all(cohort_summary$mean_ENC >= 20 & cohort_summary$mean_ENC <= 61))

p_macro <- ggplot(cohort_summary, aes(x = mean_ENC, y = mean_tAI, color = lifestyle, shape = phylum)) +
  geom_point(size = 4, alpha = 0.8) +
  labs(title = "Global Translational Selection Map (tAI vs ENC)",
       x = "Mean Effective Number of Codons (ENC)",
       y = "Mean tRNA Adaptation Index (tAI)",
       color = "Ecological Lifestyle", shape = "Taxonomic Phylum") +
  theme_bw() +
  theme(legend.position = "bottom")

# Return: An integrated scatter-marginal-boxplot assembly.
print(p_macro)

# Save high resolution image file
ggsave(file.path(sec1_dir, "translational_strategy_macro.png"), plot = p_macro, width = 10, height = 7, dpi = 300)
```

---

## Evolutionary & Cohort-Scale Analysis (Across All Species)

### Translational Strategy Mapped by Lifestyle and Taxonomy
#### Goal
Evaluate how broad macro-evolutionary descent (Phylum) and ecological pressures (Lifestyle) dictate global codon usage selection across the entire cohort landscape.

#### Methodology
Compile an integrated multi-panel visualization mapping genome-wide mean tAI against mean Effective Number of Codons (ENC/Nc). Points represent distinct species, color-coded by lifestyle categories and shaped by taxonomic phyla, paired with marginal distribution boxplots.

#### Critical Observations
Determine if specific niche adaptations (e.g., high-impact plant pathogens vs. free-living saprotrophs) cluster into separated spaces of high translational efficiency (high tAI) paired with narrowed codon usage (low ENC).

```{r translational-strategy-macro, echo=FALSE, fig.width=10, fig.height=7}
# --- Chunk: translational-strategy-macro ---
# Input: Combined cohort metric file containing mean values for every validated genome.
# Process: Create a scatter plot mapping X = mean_ENC, Y = mean_tAI. Add marginal boxplots for 'lifestyle' and 'phylum'.

# Validation: Assert that no parsed metric exceeds biological boundaries (tAI between 0 and 1; ENC between 20 and 61).
stopifnot(all(cohort_summary$mean_tAI >= 0 & cohort_summary$mean_tAI <= 1))
stopifnot(all(cohort_summary$mean_ENC >= 20 & cohort_summary$mean_ENC <= 61))

p_macro <- ggplot(cohort_summary, aes(x = mean_ENC, y = mean_tAI, color = lifestyle, shape = phylum)) +
  geom_point(size = 4, alpha = 0.8) +
  labs(title = "Global Translational Selection Map (tAI vs ENC)",
       x = "Mean Effective Number of Codons (ENC)",
       y = "Mean tRNA Adaptation Index (tAI)",
       color = "Ecological Lifestyle", shape = "Taxonomic Phylum") +
  theme_bw() +
  theme(legend.position = "bottom")

# Return: An integrated scatter-marginal-boxplot assembly.
print(p_macro)

# Save high resolution image file
ggsave(file.path(sec1_dir, "translational_strategy_macro.png"), plot = p_macro, width = 10, height = 7, dpi = 300)
```

### Selection Pressure vs Mutational Drift (The Wright's Curve)
#### Goal
Disentangle whether observed codon usage bias across species is driven by natural selection for translational speed/accuracy or by passive genomic background mutational drift.

#### Methodology
Map genome-wide average ENC against average GC3s. Overlay the theoretical mathematical Wright's curve ($H_0$: purely neutral mutational drift) defined by:
$$ENC_{\text{theoretical}} = 2 + s + \frac{29}{s^2 + (1-s)^2}$$
where $s = GC3s$.

#### Critical Observations
Species or functional cohorts pooling significantly below the theoretical line undergo active directional selection toward optimal codons. Species hugging the curve evolve primarily under neutral mutational drift.

```{r wrights-curve-analysis, echo=FALSE, fig.width=10, fig.height=7}
# --- Chunk: wrights-curve-analysis ---
# Input: Aggregated cohort matrix with 'mean_ENC' and 'mean_GC3s' vectors.
# Process: Generate a continuous math function vector tracking Wright's equation across s in [0, 1].
s_vector <- seq(0, 1, length.out = 500)
wright_theoretical <- 2 + s_vector + (29 / (s_vector^2 + (1 - s_vector)^2))
wright_curve_df <- data.frame(GC3s = s_vector, ENC = wright_theoretical)

# Plot actual species data points over the theoretical curve line.
p_wright <- ggplot() +
  geom_line(data = wright_curve_df, aes(x = GC3s, y = ENC), color = "black", size = 1.2, linetype = "solid") +
  geom_point(data = cohort_summary, aes(x = mean_GC3s, y = mean_ENC, color = lifestyle, shape = phylum), size = 3.5, alpha = 0.9) +
  ylim(20, 61) +
  labs(title = "Synonymous Selection Intensity vs. Neutral Mutation Frontier",
       x = "Synonymous Third-Position GC Content (GC3s)",
       y = "Effective Number of Codons (ENC)") +
  theme_bw()

print(p_wright)
ggsave(file.path(sec1_dir, "wrights_curve_analysis.png"), plot = p_wright, width = 10, height = 7, dpi = 300)

# Compute the perpendicular vertical distance (delta ENC) for each species as a direct proxy for selection intensity.
cohort_summary <- cohort_summary %>%
  dplyr::mutate(
    theoretical_ENC = 2 + mean_GC3s + (29 / (mean_GC3s^2 + (1 - mean_GC3s)^2)),
    delta_ENC = theoretical_ENC - mean_ENC
  )

# Validation: Flags and records any species located above the mathematical boundary line.
impossible_species <- cohort_summary %>% dplyr::filter(mean_ENC > theoretical_ENC)
if (nrow(impossible_species) > 0) {
  cat("\n**Alert: Genomic targets residing above the mutation-neutral boundary:**\n")
  print(knitr::kable(impossible_species %>% dplyr::select(sample, mean_ENC, theoretical_ENC)))
}
```


***

## Type-Level Functional Protein & Gene Adaptation (The Extremes)

### Structural Features of Protein Extremes Across Cohorts
#### Goal
Determine global amino acid and protein structural sorting rules relative to translational efficiency across the entire experimental framework.

#### Methodology
Aggregate gene sequences from all genomes into a pooled dataset. Isolate top 10% (optimized) and bottom 10% (slowed/regulatory) global tAI thresholds. Construct dodge bar charts displaying the fraction of genes containing signal peptides or transmembrane domains (TM), grouped by Phylum or Lifestyle.

#### Spatial Profiling
For the Top 10% and Bottom 10% tAI groups, generate density profile charts across 10 unified length bins tracking the positional layout of:
* Predicted protein domain boundaries (Pfam)
* Transmembrane segments (TMHMM)
* Low Complexity Regions (LCR)
This directly checks if highly optimized proteins restrict sequence complexity or transmembrane blocks to specific structural domains (e.g., N- or C-termini) compared to slower translational targets.



\```{r cross-species-structural-extremes}
# --- Chunk: cross-species-structural-extremes ---
# Input: Pooled structural metric file containing gene-by-gene measurements across all organisms ('master_gene_table.tsv').
# Process:
#   - Filter and slice global Top 10% and Bottom 10% tAI sub-cohorts.
#   - Compute percentage presence for Signal Peptides and TM domains across groups.
#   - For the spatial tracking, compute density profiles across 10 normalized alignment length bins for Pfam, TM, and LCR.
# Return: 1. Dodge bar plot of global features. 2. Continuous line charts tracing structural feature density across binned length.
# Validation: Confirm that input sequence lengths are properly normalized (0.0 to 1.0) before bin assignment.
\```

### Global Pfam & GO Terms Enrichment of Translation Extremes
#### Goal
Uncover universal metabolic or evolutionary pathways that are strictly conserved for rapid translation (Top) or translational pausing (Bottom) across taxonomic boundaries.

#### Methodology
Run hyper-geometric distribution tests for overrepresentation of Pfam domain keys and Gene Ontology (GO) terms (derived via `pfam2go` mappings) within the combined Top/Bottom 10% tAI gene pools. 

#### Critical Observations
Results are presented as profiles of top enriched functions broken down by Phylum and Lifestyle to highlight potential instances of evolutionary convergence among ecologically matching pathogens.

\```{r global-functional-enrichment}
# --- Chunk: global-functional-enrichment ---
# Input: Global sliced gene indices linked with mapped Pfam IDs and GO terms.
# Process: Execute a hyper-geometric enrichment routine (e.g., using clusterProfiler or native fisher tests). Apply Benjamini-Hochberg (BH) false discovery rate adjustments.
# Return: A consolidated significance data table containing Adjusted P-values and Enrichment Scores.
# Validation: Confirm that the background universe set matches total valid protein-coding sequences ingested by the workflow.
\```

---

## Individual Organism Deep-Dive (Per-Genome Diagnostics)
*Note: This diagnostic loop evaluates target individual assemblies dynamically or cycles sequentially through distinct species configurations.*

### Intra-Organismal tAI Distribution and Core Outliers
#### Goal
Characterize localized cell investment across individual gene tracks and identify elite, highly optimized expression targets.

#### Methodology
Render a continuous density curve mapping the internal gene-by-gene tAI distribution profile of a chosen organism. Isolate the absolute extreme top 10 and bottom 10 genes, printing them out in a tabular index accompanied by direct GO descriptions.

#### Critical Observations
Review the distribution modality. Multi-modal distribution curves reveal clear functional gene stratification, whereas highly skewed distributions indicate extreme specialization towards housekeeping genes.

\```{r single-organism-tai-distribution}
# --- Chunk: single-organism-tai-distribution ---
# Input: The standalone sample evaluation track summary ('results/per_genome/<sample>/codon_metrics/*_summary.tsv').
# Process:
#   - Generate an empirical density distribution plot for local gene tAI values.
#   - Sort the table ascending/descending to harvest the highest 10 and lowest 10 tAI gene IDs.
# Return: Combined continuous density line chart and an automated top/bottom textual index table.
# Validation: Confirm that top and bottom slices contain unique gene IDs and check for empty GO annotation strings.
\```

### Local Protein Feature Correlation
#### Goal
Examine localized structural design principles inside a single organism to check if translational speed optimization matches structural folding requirements and LCR management.

#### Methodology
Build violin/boxplot charts evaluating protein length, total TM domain count, total TM domain length, LCR counts, and absolute LCR lengths against Top 10% vs. Bottom 10% subsets.

#### N/C-Terminal LCR Polarization
Construct bar charts comparing LCR density and specific amino acid compositions across three distinct functional regions: N-terminus (0.0-0.25), Internal (0.25-0.75), and C-terminus (0.75-1.0). This verifies whether low-complexity sequences localized at the N-terminus correlate with reduced tAI to implement a controlled ribosomal translation initiation delay.

\```{r local-protein-feature-correlation}
# --- Chunk: local-protein-feature-correlation ---
# Input: Selected organism summary track integrated with structural locus adnotations.
# Process:
#   - Compile length, LCR, and TM count vectors across the extreme groups.
#   - Divide gene lengths into N-terminal, middle, and C-terminal segments and summarize internal LCR occurrence rates.
# Return: Panel displaying violin charts alongside localized structural bar plots.
# Validation: Enforce clean handling of zero values (e.g., genes entirely lacking TM domains) to prevent scale distortion.
\```

### Overlap of Structural Domains and Sequence Complexity
#### Goal
Determine if low complexity sequences (LCRs) directly collide with functional structural folds within high-efficiency translation frameworks.

#### Methodology
Calculate the physical intersection length between predicted Pfam domain bounds and localized LCR coordinates. Flag structural elements experiencing an intersection overlap of greater than 80% total length across extreme tAI brackets.

#### Critical Observations
High overlap scores in high-tAI genes point to structural optimization constraints, whereas high overlap in low-tAI genes could signify regulatory regions.

\```{r domain-lcr-structural-overlap}
# --- Chunk: domain-lcr-structural-overlap ---
# Input: Base coordinate track files mapping Pfam intervals and LCR intervals for the target genome.
# Process: Compute interval overlap lengths: (Intersection BP / Total Domain BP) * 100. Filter rows matching overlap >= 80%.
# Return: Matrix table displaying highly intersected functional domains grouped by extreme tAI status.
# Validation: Verify that coordinate index tracking systems (0-based vs 1-based) are unified across input tracks.
\```

---

## Molecular Code & Codon Efficiency Space

### Comparative Codon Relative Adaptiveness ($w_i$ Distributions)
#### Goal
Decode the underlying organismal translation rules by assessing how specific codon adaptation weights vary across lifestyles or engineering groups.

#### Methodology
Compile distribution charts of calculated relative adaptiveness weights ($w_i$) for all 61 sense codons. Contrast these profiles across ecological classifications or domains.

#### Critical Observations
Identify shifting optimal codon switches. If distinct biological groups display contrasting weights for identical codons, it highlights divergent evolution in their underlying tRNA expression profiles.

\```{r comparative-codon-weights}
# --- Chunk: comparative-codon-weights ---
# Input: Processed individual codon weight tables ('*_trna_weights.csv') compiled across genomes.
# Process: Map the 61 codon categories, sorting weights by matching amino acid clusters. Plot distributions across lifestyles.
# Return: Grid of multi-sample codon adaptiveness bar charts or density ridges.
# Validation: Ensure the normalization constraint is maintained: the preferred codon in every synonymous group must equal 1.0.
\```

### Comprehensive Codon Usage Fingerprint (RSCU Heatmap)
#### Goal
Map global codon usage preferences simultaneously across all organisms to identify evolutionary convergent trends.

#### Methodology
Construct a global two-dimensional matrix containing calculated Relative Synonymous Codon Usage (RSCU) values across all 61 sense codons for all samples. Project this matrix using a clustered heatmap with independent hierarchical tree clustering branches for rows (species) and columns (codons).

#### Critical Observations
Evaluate whether clustering topology follows strict taxonomic lineages (phylogeny) or shifts toward ecological lifestyle grouping (niche adaptation).

\```{r rscu-global-heatmap}
# --- Chunk: rscu-global-heatmap ---
# Input: Global aggregated RSCU matrix containing values for all active experimental runs.
# Process: Execute a hierarchical clustering routing using Euclidean distance metrics and complete linkage clustering.
# Return: A clustered layout heatmap plot tracking rows (species) and columns (codons).
# Validation: Confirm that row and column dimensions match the active dataset bounds (N species x 61 codons).
\```