---
title: "tAIpipe Comprehensive Analytical Report"
subtitle: "Global and Local Translation Optimization Landscape"
author: "tAIpipe Automated Framework"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: flatly
    highlight: tango
  md_document:
    variant: markdown_github
params:
  samples: ""
  dataset: ""
  per_genome_dir: ""
  output_dir: ""
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 7)

library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
library(cowplot)
library(pheatmap)
library(plotly)
library(DT)

# STANDARDIZED SECTIONS PATH DEFINITIONS
sec0_dir <- file.path(params$output_dir, "summary_report_files", "sec0_quality_control")
sec1_dir <- file.path(params$output_dir, "summary_report_files", "sec1_evolutionary_cohort")
sec2_dir <- file.path(params$output_dir, "summary_report_files", "sec2_cross_species_functional")
sec3_dir <- file.path(params$output_dir, "summary_report_files", "sec3_individual_deep_dive")
sec4_dir <- file.path(params$output_dir, "summary_report_files", "sec4_codon_weights_weights")

# Ensure asset directories exist
for (dir_path in list(sec0_dir, sec1_dir, sec2_dir, sec3_dir, sec4_dir)) {
  if (!dir.exists(dir_path)) dir.create(dir_path, recursive = TRUE)
}

# LOAD INPUT DATASETS
# Dataset contains phenotypic/taxonomic information per sample
dataset_df <- read_delim(params$dataset, delim = "\t", show_col_types = FALSE) %>%
  filter(include == TRUE | include == "True")

# Samples contains aggregated per-gene coding metrics and annotations
samples_df <- read_delim(params$samples, delim = "\t", show_col_types = FALSE)

# Combined analytical view for downstream chunks
analysis_df <- samples_df %>%
  inner_join(dataset_df, by = "sample")
```

# tAIpipe Analytical Report: Global and Local Translation Landscape

***

## Quality Control & Dataset Auditing

### Cohort Ingestion Summary
#### Goal
Verify data ingestion and experiment layout integrity before computational evaluation. Ensure taxonomic and ecological categories are represented.

#### Methodology
Compile a cohort overview table summarizing total counts of independent operational taxonomic units (samples) across Domains, Kingdoms, Phyla, and Lifestyles.

#### Critical Observations
Look for extreme class imbalances or sample omissions. If a specific lifestyle or phylum contains too few records compared to the rest of the cohort, a text warning must alert the user to potential statistical inflation during downstream grouping.

```{r cohort-ingestion-summary, #TODO }
# --- Chunk: cohort-ingestion-summary ---
# Input: Phenotypic and taxonomic metadata table (dataset_df)
# Process: Aggregate counts across taxonomic hierarchies and lifestyle classifications.

cohort_table <- dataset_df %>%
  group_by(domain, kingdom, phylum, lifestyle) %>%
  summarise(sample_count = n(), .groups = 'drop')

datatable(cohort_table, 
          options = list(pageLength = 10, dom = 'tip'),
          rownames = FALSE,
          caption = "Table 1: Overview of ingested biological samples stratified by taxonomy and lifestyle niche.")
```



### Organism-Level Assembly & Prediction QC
#### Goal
Identify analytical outliers, structural sequence assembly fragmentation artifacts, and potential downstream tRNA prediction errors.

#### Methodology
Generate a dual panel of sample-wide scatter plots:
1. Genome Assembly Size (Mbp) vs. Total Predicted tRNA Gene Count.
2. Genome-wide Average GC Content vs. Genome-wide Average Synonymous Third-Position GC Content (GC3s).

#### Critical Observations
Legitimate biological trends exhibit linear scaling between genome size and tRNA pools. Mark severe technical outliers in red with text labels. For instance, a bloated genome sequence with an artificially suppressed tRNA count indicates low assembly coverage or raw file contamination.

#### Notes
Samples identified as suspicious through the entire pipeline auditing process (across relevant data splits) must be explicitly flagged and listed here.

```{r assembly-prediction-qc-plots, echo=FALSE, fig.width=12, fig.height=6}
# --- Chunk: assembly-prediction-qc-plots ---
# Input: Comprehensive analysis dataset mapping per-gene observations and genomic patterns.

cohort_summary <- analysis_df %>%
  group_by(sample, species, phylum, lifestyle, domain) %>%
  summarise(
    mean_tAI = mean(tAI, na.rm = TRUE),
    mean_ENC = mean(ENC, na.rm = TRUE),
    mean_GC = mean(GC, na.rm = TRUE),
    mean_GC3s = mean(GC3s, na.rm = TRUE),
    total_genes = n(),
    .groups = 'drop'
  ) %>%
  mutate(total_tRNA_count = sapply(sample, function(s) {
    p <- file.path(params$per_genome_dir, s, "trnascan", paste0(s, "_trnascan.tsv"))
    if(file.exists(p)) nrow(read_tsv(p, show_col_types = FALSE)) else 0
  })) %>%
  mutate(genome_size_Mbp = total_genes * 0.003) 

# Cook's distance filter to identify structural outliers
fit <- lm(total_tRNA_count ~ genome_size_Mbp, data = cohort_summary)
cohort_summary$cooksd <- cooks.distance(fit)
n_samples <- nrow(cohort_summary)
cohort_summary <- cohort_summary %>%
  mutate(Outlier = ifelse(cooksd > (4 / n_samples), "Flagged Outlier", "Normal Lineage"))

# Interactive Plot 1: Estimated Genome Size vs tRNA Count
p1 <- ggplot(cohort_summary, aes(x = genome_size_Mbp, y = total_tRNA_count, text = paste("Sample:", sample, "<br>Species:", species))) +
  geom_point(aes(color = Outlier), size = 3) +
  geom_smooth(method = "lm", color = "black", se = FALSE, linetype = "dashed") +
  scale_color_manual(values = c("Flagged Outlier" = "red", "Normal Lineage" = "black")) +
  labs(title = "tRNA Repertoire Scaling Analysis", x = "Estimated Genome Size (Mbp)", y = "Total tRNA Gene Count") +
  theme_bw()

# Interactive Plot 2: GC vs GC3s Concordance
p2 <- ggplot(cohort_summary, aes(x = mean_GC, y = mean_GC3s, color = lifestyle, text = paste("Sample:", sample, "<br>Species:", species))) +
  geom_point(size = 3) +
  labs(title = "Nucleotide Composition Concordance", x = "Genome Mean GC", y = "Genome Mean GC3s") +
  theme_bw()

# Export static versions for asset caching
ggsave(file.path(sec0_dir, "trna_repertoire_scaling.png"), plot = p1, width = 6, height = 5)
ggsave(file.path(sec0_dir, "nucleotide_composition_concordance.png"), plot = p2, width = 6, height = 5)

# Render interactive subplots
subplot(ggplotly(p1, tooltip = "text"), ggplotly(p2, tooltip = "text"), nrows = 1, titleX = TRUE, titleY = TRUE)
```



# ```{r translational-strategy-macro, echo=FALSE, fig.width=10, fig.height=7}
# # --- Chunk: translational-strategy-macro ---
# # Input: Compiled cohort summary data frame from the previous checkpoint.
# # Validation: Enforce absolute biological boundaries before evaluation loops.
# stopifnot(all(cohort_summary$mean_tAI >= 0 & cohort_summary$mean_tAI <= 1))
# stopifnot(all(cohort_summary$mean_ENC >= 20 & cohort_summary$mean_ENC <= 61))

# p_macro <- ggplot(cohort_summary, aes(x = mean_ENC, y = mean_tAI, color = lifestyle, shape = phylum,
#                                      text = paste("Sample:", sample, "<br>Species:", species))) +
#   geom_point(size = 4, alpha = 0.8) +
#   labs(title = "Global Translational Selection Map (tAI vs ENC)",
#        x = "Mean Effective Number of Codons (ENC)",
#        y = "Mean tRNA Adaptation Index (tAI)",
#        color = "Ecological Lifestyle", shape = "Taxonomic Phylum") +
#   theme_bw()

# # Save static artifact
# ggsave(file.path(sec1_dir, "translational_strategy_macro.png"), plot = p_macro, width = 10, height = 7, dpi = 300)

# ggplotly(p_macro, tooltip = "text")
# ```

***

## Evolutionary & Cohort-Scale Analysis (Across All Species)

### Translational Strategy Mapped by Lifestyle and Taxonomy
#### Goal
Evaluate how broad macro-evolutionary descent (Phylum) and ecological pressures (Lifestyle) dictate global codon usage selection across the entire cohort landscape.

#### Methodology
Compile an integrated multi-panel visualization mapping genome-wide mean tAI against mean Effective Number of Codons (ENC/Nc). Points represent distinct species, color-coded by lifestyle categories and shaped by taxonomic phyla, paired with marginal distribution boxplots.

#### Critical Observations
Determine if specific niche adaptations (e.g., high-impact plant pathogens vs. free-living saprotrophs) cluster into separated spaces of high translational efficiency (high tAI) paired with narrowed codon usage (low ENC).

```{r translational-strategy-macro, echo=FALSE, fig.width=10, fig.height=7}
# --- Chunk: translational-strategy-macro ---
# Input: Compiled cohort summary data frame from the previous checkpoint.
# Validation: Enforce absolute biological boundaries before evaluation loops.
stopifnot(all(cohort_summary$mean_tAI >= 0 & cohort_summary$mean_tAI <= 1))
stopifnot(all(cohort_summary$mean_ENC >= 20 & cohort_summary$mean_ENC <= 61))

p_macro <- ggplot(cohort_summary, aes(x = mean_ENC, y = mean_tAI, color = lifestyle, shape = phylum,
                                     text = paste("Sample:", sample, "<br>Species:", species))) +
  geom_point(size = 4, alpha = 0.8) +
  labs(title = "Global Translational Selection Map (tAI vs ENC)",
       x = "Mean Effective Number of Codons (ENC)",
       y = "Mean tRNA Adaptation Index (tAI)",
       color = "Ecological Lifestyle", shape = "Taxonomic Phylum") +
  theme_bw()

# Save static artifact
ggsave(file.path(sec1_dir, "translational_strategy_macro.png"), plot = p_macro, width = 10, height = 7, dpi = 300)

ggplotly(p_macro, tooltip = "text")
```

### Selection Pressure vs Mutational Drift (The Wright's Curve)
#### Goal
Disentangle whether observed codon usage bias across species is driven by natural selection for translational speed/accuracy or by passive genomic background mutational drift.

#### Methodology
Map genome-wide average ENC against average GC3s. Overlay the theoretical mathematical Wright's curve ($H_0$: purely neutral mutational drift) defined by:
$$ENC_{\text{theoretical}} = 2 + s + \frac{29}{s^2 + (1-s)^2}$$
where $s = GC3s$.

#### Critical Observations
Species or functional cohorts pooling significantly below the theoretical line undergo active directional selection toward optimal codons. Species hugging the curve evolve primarily under neutral mutational drift.


```{r wrights-curve-analysis, echo=FALSE, fig.width=10, fig.height=7}
# --- Chunk: wrights-curve-analysis ---
# Input: Aggregated cohort matrix with 'mean_ENC' and 'mean_GC3s' vectors.
# Process: Generate a continuous math function vector tracking Wright's equation across s in [0, 1].
s_vector <- seq(0, 1, length.out = 500)
wright_theoretical <- 2 + s_vector + (29 / (s_vector^2 + (1 - s_vector)^2))
wright_curve_df <- data.frame(GC3s = s_vector, ENC = wright_theoretical)

# Plot actual species data points over the theoretical curve line.
p_wright <- ggplot() +
  geom_line(data = wright_curve_df, aes(x = GC3s, y = ENC), color = "black", size = 1.2, linetype = "solid") +
  geom_point(data = cohort_summary, aes(x = mean_GC3s, y = mean_ENC, color = lifestyle, shape = phylum), size = 3.5, alpha = 0.9) +
  ylim(20, 61) +
  labs(title = "Synonymous Selection Intensity vs. Neutral Mutation Frontier",
       x = "Synonymous Third-Position GC Content (GC3s)",
       y = "Effective Number of Codons (ENC)") +
  theme_bw()

print(p_wright)
ggsave(file.path(sec1_dir, "wrights_curve_analysis.png"), plot = p_wright, width = 10, height = 7, dpi = 300)

# Compute the perpendicular vertical distance (delta ENC) for each species as a direct proxy for selection intensity.
cohort_summary <- cohort_summary %>%
  dplyr::mutate(
    theoretical_ENC = 2 + mean_GC3s + (29 / (mean_GC3s^2 + (1 - mean_GC3s)^2)),
    delta_ENC = theoretical_ENC - mean_ENC
  )

# Validation: Flags and records any species located above the mathematical boundary line.
impossible_species <- cohort_summary %>% dplyr::filter(mean_ENC > theoretical_ENC)
if (nrow(impossible_species) > 0) {
  cat("\n**Alert: Genomic targets residing above the mutation-neutral boundary:**\n")
  print(knitr::kable(impossible_species %>% dplyr::select(sample, mean_ENC, theoretical_ENC)))
}
```

***

## Cross-Species Functional Protein & Gene Adaptation (The Extremes)

### Global Pfam & GO Terms Enrichment of Translation Extremes
#### Goal
Uncover universal metabolic or evolutionary pathways that are strictly conserved for rapid translation (Top) or translational pausing (Bottom) across taxonomic boundaries.

#### Methodology
Run hyper-geometric distribution tests for overrepresentation of Pfam domain keys and Gene Ontology (GO) terms (derived via `pfam2go` mappings) within the combined Top/Bottom 10% tAI gene pools. 

#### Critical Observations
Results are presented as profiles of top enriched functions broken down by Phylum and Lifestyle to highlight potential instances of evolutionary convergence among ecologically matching pathogens.

```{r global-functional-enrichment}
# --- Chunk: global-functional-enrichment ---
# Input: Combined multi-sample dataframe containing gene information and go_terms annotation fields.

# Slicing out top and bottom tails from the analytical workspace
enriched_data <- analysis_df %>%
  group_by(sample) %>%
  mutate(
    rank_tai = percent_rank(tAI),
    Cohort_Split = case_when(
      rank_tai >= 0.90 ~ "Top 10% Optimized",
      rank_tai <= 0.10 ~ "Bottom 10% Suppressed",
      TRUE ~ "Background"
    )
  ) %>%
  filter(Cohort_Split != "Background" & !is.na(go_terms) & go_terms != "") %>%
  separate_rows(go_terms, sep = ";") %>%
  group_by(lifestyle, Cohort_Split, go_terms) %>%
  summarise(Observed_Count = n(), .groups = 'drop') %>%
  arrange(desc(Observed_Count))

datatable(enriched_data, 
          options = list(pageLength = 10, searchHighlight = TRUE),
          rownames = FALSE,
          caption = "Table 2: Gene Ontology (GO) terms representation count across functional tails.")
```

***

## Individual Organism Deep-Dive (Per-Genome Diagnostics)
*Note: This diagnostic loop evaluates target individual assemblies dynamically or cycles sequentially through distinct species configurations.*

### Intra-Organismal tAI Distribution and Core Outliers
#### Goal
Characterize localized cell investment across individual gene tracks and identify elite, highly optimized expression targets.

#### Methodology
Render a continuous density curve mapping the internal gene-by-gene tAI distribution profile of a chosen organism. Isolate the absolute extreme top 10 and bottom 10 genes, printing them out in a tabular index accompanied by direct GO descriptions.

#### Critical Observations
Review the distribution modality. Multi-modal distribution curves reveal clear functional gene stratification, whereas highly skewed distributions indicate extreme specialization towards housekeeping genes.

```{r single-organism-tai-distribution}
# --- Chunk: single-organism-tai-distribution ---
# Input: Per-gene analytics dataset containing expression indices.

p_density <- ggplot(analysis_df, aes(x = tAI, fill = sample)) +
  geom_density(alpha = 0.4) +
  labs(title = "Intra-Organismal tAI Density Distribution Profiles",
       x = "tRNA Adaptation Index (tAI)", y = "Probability Density Function") +
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave(file.path(sec3_dir, "intra_organismal_tai_density.png"), plot = p_density, width = 10, height = 6)

ggplotly(p_density)

# Slice and print out extreme candidate arrays for review
extreme_genes_table <- analysis_df %>%
  group_by(sample) %>%
  mutate(Rank = min_rank(desc(tAI))) %>%
  filter(Rank <= 10) %>%
  select(sample, gene_id, protein_id, tAI, ENC, protein_name, go_terms) %>%
  arrange(sample, desc(tAI))

datatable(extreme_genes_table, 
          options = list(pageLength = 10, scrollX = TRUE), 
          rownames = FALSE,
          caption = "Table 3: High-throughput identification of elite highly optimized genes (Top 10 per Sample).")
```


***

## Molecular Code & Codon Efficiency Space

### Comparative Codon Relative Adaptiveness ($w_i$ Distributions)
#### Goal
Decode the underlying organismal translation rules by assessing how specific codon adaptation weights vary across lifestyles or engineering groups.

#### Methodology
Compile distribution charts of calculated relative adaptiveness weights ($w_i$) for all 61 sense codons. Contrast these profiles across ecological classifications or domains.

#### Critical Observations
Identify shifting optimal codon switches. If distinct biological groups display contrasting weights for identical codons, it highlights divergent evolution in their underlying tRNA expression profiles.

```{r comparative-codon-weights}
# --- Chunk: comparative-codon-weights ---
# Input: Multi-sample loop pulling local tRNA relative adaptation values.

weights_list <- list()
for(s_id in unique(analysis_df$sample)) {
  w_path <- file.path(params$per_genome_dir, s_id, "codon_metrics", paste0(s_id, "_tai.csv"))
  if(file.exists(w_path)) {
    # Expected layout: parsed relative weights index sheet
    w_data <- read_csv(w_path, show_col_types = FALSE)
    if("codon" %in% colnames(w_data) & "weight" %in% colnames(w_data)) {
      w_data$sample <- s_id
      weights_list[[s_id]] <- w_data
    }
  }
}

if(length(weights_list) > 0) {
  weights_df <- bind_rows(weights_list) %>% inner_join(dataset_df, by = "sample")
  
  p_weights <- ggplot(weights_df, aes(x = codon, y = weight, fill = lifestyle)) +
    geom_boxplot(outlier.size = 0.5) +
    labs(title = "Codon Relative Adaptiveness Distribution Matrix Across Lifestyles",
         x = "Codon Triplet Category", y = "Relative Adaptive Weight (wi)") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1), legend.position = "bottom")
  
  ggsave(file.path(sec4_dir, "comparative_codon_weights_distribution.png"), plot = p_weights, width = 12, height = 6)
  ggplotly(p_weights)
} else {
  cat("\n*No compiled codon relative adaptiveness metrics files found across active workspace configurations.*\n")
}
```

### Comprehensive Codon Usage Fingerprint (RSCU Heatmap)
#### Goal
Map global codon usage preferences simultaneously across all organisms to identify evolutionary convergent trends.

#### Methodology
Construct a global two-dimensional matrix containing calculated Relative Synonymous Codon Usage (RSCU) values across all 61 sense codons for all samples. Project this matrix using a clustered heatmap with independent hierarchical tree clustering branches for rows (species) and columns (codons).

#### Critical Observations
Evaluate whether clustering topology follows strict taxonomic lineages (phylogeny) or shifts toward ecological lifestyle grouping (niche adaptation).

```{r rscu-global-heatmap}
# --- Chunk: rscu-global-heatmap ---
# Input: Iterative retrieval loop across computed RSCU profiling results.

rscu_list <- list()
for (i in 1:nrow(dataset_df)) {
  s_id <- dataset_df$sample[i]
  rscu_path <- file.path(params$per_genome_dir, s_id, "codon_metrics", paste0(s_id, "_rscu.csv"))
  
  if (file.exists(rscu_path)) {
    rscu_data <- read_csv(rscu_path, show_col_types = FALSE)
    rscu_mean <- rscu_data %>% summarise(across(where(is.numeric), mean, na.rm = TRUE))
    rscu_mean$sample <- s_id
    rscu_list[[s_id]] <- rscu_mean
  }
}

if (length(rscu_list) > 0) {
  rscu_matrix_df <- bind_rows(rscu_list)
  rownames_vectors <- rscu_matrix_df$sample
  
  rscu_matrix <- rscu_matrix_df %>% select(-sample) %>% as.matrix()
  rownames(rscu_matrix) <- rownames_vectors
  rscu_matrix <- rscu_matrix[, colSums(is.na(rscu_matrix)) < nrow(rscu_matrix)]
  
  heatmap_file <- file.path(sec1_dir, "rscu_global_heatmap.png")
  
  pheatmap(rscu_matrix, 
           clustering_distance_rows = "euclidean",
           clustering_distance_cols = "euclidean",
           clustering_method = "complete",
           filename = heatmap_file,
           width = 10, height = 8,
           main = "Global Codon Usage Fingerprint Clustering Matrix (RSCU)")
  
  knitr::include_graphics(heatmap_file)
} else {
  message("RSCU metric matrix data arrays missing across current execution partitions.")
}
```